---
title: "CIS 690 GDM Prediction Project"
author: "Group Members: Abenezer Ghiday, Daniela Ortega, Davis Reinhard, Manav Mendonca"
date: "Jun, 2026"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# 2a: Load and Inspect

```{r data-inspection}
# Load the dataset
patients <- read.csv("patients.csv")

# View first few rows
head(patients)

# Check structure and dimensions
str(patients)
dim(patients)
```

# 2b: Summary Statistics

```{r summary-stats}
# Summary statistics for all variables
summary(patients)

# Standard deviation for each variable
sapply(patients, sd)
```
##Interpretation-
Pregnancies: Mean = 3.85, Median = 3, SD = 3.37, Q1 = 1, Q3 = 6, Range = 0 to 17. The distribution is right skewed given the gap between the mean and median. Zero pregnancies is a valid value here since it simply means no prior pregnancies.

Glucose: Mean = 120.9, Median = 117, SD = 31.97, Q1 = 99, Q3 = 140.25, Range = 0 to 199. The mean and median are close together suggesting a fairly symmetric distribution. The minimum of zero is not clinically possible and will be treated as missing.

BloodPressure: Mean = 69.11, Median = 72, SD = 19.36, Q1 = 62, Q3 = 80, Range = 0 to 122. Values are centered around a diastolic range of 62 to 80. The minimum of zero is not a valid blood pressure reading and will be treated as missing.

SkinThickness: Mean = 20.54, Median = 23, SD = 15.95, Q1 = 0, Q3 = 32, Range = 0 to 99. The first quartile being zero tells us that at least 25% of the values are zero, which confirms the high level of missingness at 29.6% of the data.

Insulin: Mean = 79.8, Median = 30.5, SD = 115.24, Q1 = 0, Q3 = 127.25, Range = 0 to 846. The large gap between mean and median and the extremely high SD indicate a heavily right skewed distribution with extreme outliers. The first quartile of zero confirms that more than 25% of values are zero, making this the most problematic variable with 48.7% missing.

BMI: Mean = 31.99, Median = 32, SD = 7.88, Q1 = 27.3, Q3 = 36.6, Range = 0 to 67.1. The mean and median are nearly identical suggesting a symmetric distribution. The minimum of zero is not clinically possible and will be treated as missing.

Pedigree: Mean = 0.47, Median = 0.37, SD = 0.33, Q1 = 0.24, Q3 = 0.63, Range = 0.08 to 2.42. Right skewed with most values concentrated below 1.0. No zeros or missing values.

Age: Mean = 33.24, Median = 29, SD = 11.76, Q1 = 24, Q3 = 41, Range = 21 to 81. Right skewed with most patients between their mid-twenties and early forties. No zeros or missing values. Given that this dataset is centered around GDM, 81 as the max age is clinically unusual and should be interpreted cautiously.

Diagnosis: Mean = 0.349 confirms the class distribution with approximately 35% of patients having a GDM diagnosis and 65% not.

The most immediate concern from these statistics is the zero values in Glucose, BloodPressure, SkinThickness, Insulin, and BMI. Treating these implausible clinical values as missing or invalid is consistent with diabetes prediction preprocessing practices [16]. Insulin is the most problematic with a first quartile of zero, a mean far above the median, and an SD of 115 indicating extreme variability. These zeros are not real measurements and will be addressed in Step 2d.

# 2c: Visualization, used full raw data for exploration

```{r}
library(ggplot2)
```

## Plot 1: Class Distribution

```{r plot1}
ggplot(patients, aes(x = factor(Diagnosis), fill = factor(Diagnosis))) +
  geom_bar(width = 0.5) +
  scale_x_discrete(labels = c("0" = "No GDM", "1" = "GDM")) +
  scale_fill_manual(values = c("steelblue", "tomato"),
                    labels = c("No GDM", "GDM"),
                    name   = "Diagnosis") +
  labs(title = "Class Distribution: GDM vs No GDM",
       x = "Diagnosis", y = "Count") +
  theme_minimal()
```
##Interpretation-
The bar chart displays the number of patients in each diagnosis category. The No GDM group contains approximately 500 (~65%) patients, while the GDM group contains approximately 268 (~35%) patients. This indicates that the dataset contains more patients without gestational diabetes than patients with gestational diabetes, suggesting a moderate class imbalance that should be considered during model development and evaluation. The visualization was created using the raw dataset before any preprocessing or balancing techniques were applied.

## Plot 2: Glucose Distribution by Diagnosis

```{r }
ggplot(patients, aes(x = Glucose, fill = factor(Diagnosis))) +
  geom_histogram(bins = 30, alpha = 0.6, position = "identity") +
  scale_fill_manual(values = c("steelblue", "tomato"),
                    labels = c("No GDM", "GDM"),
                    name   = "Diagnosis") +
  labs(title = "Glucose Distribution by Diagnosis",
       x = "Glucose (mg/dL)", y = "Count") +
  theme_minimal()
```
##Interpretation-
The histogram shows that patients diagnosed with GDM generally have higher glucose levels than patients in the No GDM group. The GDM distribution is shifted to the right, with a larger concentration of observations above 140 mg/dL, while the No GDM group is concentrated around lower glucose values. Although there is some overlap between the two groups, the separation between the distributions suggests that glucose is likely an important predictor of gestational diabetes in this dataset. This aligns with our literature review, where glucose measurements were considered as consistently supported predictors in GDM prediction models [4], [5]. The visualization also reveals a few glucose values of zero, which are not clinically realistic and will be treated as missing values during data preparation.

## Plot 3: BMI Distribution by Diagnosis

```{r BMI-by-diagnosis}
ggplot(patients, aes(x = BMI, fill = factor(Diagnosis))) +
  geom_histogram(bins = 30, alpha = 0.6, position = "identity") +
  scale_fill_manual(values = c("steelblue", "tomato"),
                    labels = c("No GDM", "GDM"),
                    name   = "Diagnosis") +
  labs(title = "BMI Distribution by Diagnosis",
       x = "BMI", y = "Count") +
  theme_minimal()
```
##Interpretation-
The histogram shows that patients diagnosed with GDM tend to have higher BMI values than patients in the No GDM group. The GDM distribution is shifted slightly to the right, indicating that higher BMI may be associated with an increased risk of gestational diabetes. This is also consistent with the literature, where BMI IS identified as one of the most commonly used and supported predictors in GDM prediction models [4], [9]. However, there is considerable overlap between the two groups, suggesting that BMI alone is not sufficient to distinguish between GDM and No GDM cases. The visualization also shows a few BMI values of zero, which are not clinically realistic and will be treated as missing values during data preparation.

## Plot 4: Boxplots of Key Predictors by Diagnosis

```{r boxplots-key-predictors}
ggplot(patients, aes(x = factor(Diagnosis), y = Glucose,
                     fill = factor(Diagnosis))) +
  geom_boxplot(alpha = 0.7) +
  scale_x_discrete(labels = c("0" = "No GDM", "1" = "GDM")) +
  scale_fill_manual(values = c("steelblue", "tomato"),
                    name   = "Diagnosis") +
  labs(title = "Glucose by Diagnosis",
       x = "Diagnosis", y = "Glucose (mg/dL)") +
  theme_minimal()

ggplot(patients, aes(x = factor(Diagnosis), y = BMI,
                     fill = factor(Diagnosis))) +
  geom_boxplot(alpha = 0.7) +
  scale_x_discrete(labels = c("0" = "No GDM", "1" = "GDM")) +
  scale_fill_manual(values = c("steelblue", "tomato"),
                    name   = "Diagnosis") +
  labs(title = "BMI by Diagnosis",
       x = "Diagnosis", y = "BMI") +
  theme_minimal()

ggplot(patients, aes(x = factor(Diagnosis), y = Age,
                     fill = factor(Diagnosis))) +
  geom_boxplot(alpha = 0.7) +
  scale_x_discrete(labels = c("0" = "No GDM", "1" = "GDM")) +
  scale_fill_manual(values = c("steelblue", "tomato"),
                    name   = "Diagnosis") +
  labs(title = "Age by Diagnosis",
       x = "Diagnosis", y = "Age (years)") +
  theme_minimal()

ggplot(patients, aes(x = factor(Diagnosis), y = Pedigree,
                     fill = factor(Diagnosis))) +
  geom_boxplot(alpha = 0.7) +
  scale_x_discrete(labels = c("0" = "No GDM", "1" = "GDM")) +
  scale_fill_manual(values = c("steelblue", "tomato"),
                    name   = "Diagnosis") +
  labs(title = "Pedigree by Diagnosis",
       x = "Diagnosis", y = "Pedigree") +
  theme_minimal()
```
##Interpretation-
GDM patients show higher median values for Glucose, BMI, Age, and Pedigree compared to No GDM patients. Glucose shows the clearest separation between groups with the GDM median noticeably higher. BMI and Age also show differences but with more overlap. Pedigree, which represents family history related diabetes risk, is also higher in the GDM group, consistent with Zhang et al. [4] and Lee et al. [9] who identified family history as a key predictor.

## Plot 5: Age vs Glucose Scatterplot by Diagnosis

```{r age-glucose-scatter}
ggplot(patients, aes(x = Age, y = Glucose,
                     color = factor(Diagnosis))) +
  geom_point(alpha = 0.6) +
  scale_color_manual(values = c("steelblue", "tomato"),
                     labels = c("No GDM", "GDM"),
                     name   = "Diagnosis") +
  labs(title = "Age vs Glucose by Diagnosis",
       x = "Age (years)", y = "Glucose (mg/dL)") +
  theme_minimal()
```
##Interpretation-
The scatterplot shows the relationship between age and glucose levels for patients with and without gestational diabetes. Patients diagnosed with GDM tend to have higher glucose values across most age groups, while patients without GDM are more concentrated at lower glucose levels. There does not appear to be a strong linear relationship between age and glucose, but glucose levels seem to be more useful than age for distinguishing between the two diagnosis groups. A few glucose values of zero are also present and will be treated as missing values during data preparation. This plot also shows a small number of unusually high ages for the GDM dataset, so those observations should be interpreted cautiously. 

# 2d: Missing Value Analysis

## Step 1: Check for NA values and Zero values

```{r na-zero-check}
# Check for actual NA values first
colSums(is.na(patients))

# Check zero counts per variable
colSums(patients == 0)
```
##Interpretation-
No variables contain NA values. However Glucose, BloodPressure, SkinThickness, Insulin, and BMI contain zeros that are not physiologically possible. These are treated as missing data. Again, this is consistent with diabetes prediction preprocessing practices where biologically implausible zero values are treated as missing or invalid rather than true measurements[16]. Pregnancies and Diagnosis zeros are valid and left unchanged. Pedigree and Age have no zeros and no missing values.

## Step 2: Replace invalid zeros with NA

```{r replace-zeros}
# Replace zeros with NA only in variables where zero is not valid
patients_clean    <- patients
invalid_zero_vars <- c("Glucose", "BloodPressure", "SkinThickness",
                       "Insulin", "BMI")

patients_clean[invalid_zero_vars] <- lapply(
  patients_clean[invalid_zero_vars],
  function(x) ifelse(x == 0, NA, x)
)

# Confirm missing values after replacement
colSums(is.na(patients_clean))
```
##Interpretation-
After replacing invalid zeros with NA, five variables now show missing values. Insulin has the highest missingness with 374 missing values followed by SkinThickness with 227. BloodPressure has 35 missing values, BMI has 11, and Glucose has only 5. Pregnancies and Diagnosis zeros were left unchanged since they represent valid clinical values. Pedigree and Age remain complete with no missing values.

## Step 3: Split the data BEFORE imputation to prevent data leakage

```{r data-split}
# Split must happen before imputation
# Any median calculated after splitting uses training data only
# This prevents test data from influencing the imputation process
# Split the data 70/30
# 70% training gives the model enough data to learn from
# 30% testing gives a more reliable evaluation with this small dataset

set.seed(123)

n         <- nrow(patients_clean)
train_idx <- sample(seq_len(n), size = floor(0.70 * n))

train_data <- patients_clean[ train_idx, ]
test_data  <- patients_clean[-train_idx, ]

cat("Training set:", nrow(train_data), "observations\n")
cat("Testing set: ", nrow(test_data),  "observations\n")

cat("\nTraining class distribution:\n")
table(train_data$Diagnosis)

cat("\nTesting class distribution:\n")
table(test_data$Diagnosis)

# Convert Diagnosis to numeric to allow correlation analysis later
train_data$Diagnosis <- as.numeric(as.character(train_data$Diagnosis))
test_data$Diagnosis  <- as.numeric(as.character(test_data$Diagnosis))
```
##Interpretation-
The dataset was split into 70% training (537 observations) and 30% testing (231 observations). A 70/30 split was chosen over 80/20 because the dataset only contains 768 observations. With 80/20 the test set would have contained only around 54 positive GDM cases which is too small to produce reliable sensitivity and recall estimates. The 70/30 split gives 81 positive GDM cases in the test set which provides a more stable evaluation of model performance on unseen data. According to Sivakumar, et. al a 70/30 split is widely accepted across different fields for its robustness and providing enough data for both training and testing.
The class distribution is well preserved in both sets with roughly 65% No GDM and 35% GDM in both training and testing, confirming the split was done correctly.

## Step 4: Calculate medians from training data only

```{r median-imputation-on-training-split}
# Calculate median from training data only
# na.rm = TRUE ensures the invalid zeros already replaced with NA
# are excluded so the median only uses real observed clinical values

glucose_median <- median(train_data$Glucose,       na.rm = TRUE)
bp_median      <- median(train_data$BloodPressure, na.rm = TRUE)
bmi_median     <- median(train_data$BMI,           na.rm = TRUE)
skin_median    <- median(train_data$SkinThickness, na.rm = TRUE)
insulin_median <- median(train_data$Insulin,       na.rm = TRUE)

cat("Training medians used for imputation:\n")
cat("Glucose:       ", glucose_median,  "\n")
cat("BloodPressure: ", bp_median,       "\n")
cat("BMI:           ", bmi_median,      "\n")
cat("SkinThickness: ", skin_median,     "\n")
cat("Insulin:       ", insulin_median,  "\n")
```
##Interpretation-
Median values were calculated from the training data only. The medians used for imputation were 119 mg/dL for Glucose, 72 mmHg for BloodPressure, 32 for BMI, 29 mm for SkinThickness, and 125 mu U/ml for Insulin. These medians were then applied to both the training and test sets to fill in the missing values. Using training medians for the test set ensures no information from the test data influences the imputation process, which prevents data leakage as highlighted by Netayawijit et al. [16].

Based on the literature, we found the most useful predictors to be glucose measurements, BMI, age, family history of diabetes. [4],[8],[9]. Given that Skinthickness and BMI, which are the features with most missigness present, do not fall under this category we felt confident utilizing median imputation. Berkelam, et.al found that median imputation was noniferior to other more complex imputation methods in cardiovascular risk prediction, "provided that most important predictor variables are present", which in our case had very low missingness. However, it is important to mention it's limitation such as the assumption of "missigness being unrelated to unseen variables" [Benhamza, et.al]

## Step 5: Apply those same medians to both training and test data

```{r impute}
# Apply imputation to training data first
train_data$Glucose <- ifelse(is.na(train_data$Glucose),
                             glucose_median, train_data$Glucose)

train_data$BloodPressure <- ifelse(is.na(train_data$BloodPressure),
                                   bp_median, train_data$BloodPressure)

train_data$BMI <- ifelse(is.na(train_data$BMI),
                         bmi_median, train_data$BMI)

train_data$SkinThickness <- ifelse(is.na(train_data$SkinThickness),
                                   skin_median, train_data$SkinThickness)

train_data$Insulin <- ifelse(is.na(train_data$Insulin),
                             insulin_median, train_data$Insulin)

# Apply the exact same training medians to test data
# The test data never influences what values are used
test_data$Glucose <- ifelse(is.na(test_data$Glucose),
                            glucose_median, test_data$Glucose)

test_data$BloodPressure <- ifelse(is.na(test_data$BloodPressure),
                                  bp_median, test_data$BloodPressure)

test_data$BMI <- ifelse(is.na(test_data$BMI),
                        bmi_median, test_data$BMI)

test_data$SkinThickness <- ifelse(is.na(test_data$SkinThickness),
                                  skin_median, test_data$SkinThickness)

test_data$Insulin <- ifelse(is.na(test_data$Insulin),
                            insulin_median, test_data$Insulin)

# Confirm no missing values remain in either set
cat("Missing in training after imputation:\n")
colSums(is.na(train_data))

cat("\nMissing in test after imputation:\n")
colSums(is.na(test_data))
```
##Interpretation-
After imputation both the training and test sets now have zero missing values across all nine variables. The medians were calculated from the training set only and then applied to both sets. This is important because if the medians had been calculated from the full dataset before splitting, the test data would have quietly influenced the imputation values which is a form of data leakage. By splitting first and imputing second the test set stays completely untouched and gives an honest picture of how the model will actually perform on new patients.

# 2e: Outlier Detection and Handling
## Detect Outliers

```{r iqr}
# Function to count outliers using IQR rule
detect_outliers <- function(x) {
  Q1  <- quantile(x, 0.25, na.rm = TRUE)
  Q3  <- quantile(x, 0.75, na.rm = TRUE)
  IQR <- Q3 - Q1
  sum(x < (Q1 - 1.5 * IQR) | x > (Q3 + 1.5 * IQR), na.rm = TRUE)
}

# Count outliers in training data
cat("Outlier counts in training data:\n")
sapply(train_data, detect_outliers)
```
##Interpretation-
The IQR rule identified outliers in seven of the nine variables. Insulin has the most with 237 outliers, which is consistent with its heavily skewed distribution and extreme values seen in the summary statistics. SkinThickness follows with 39 outliers. Glucose and Diagnosis show no outliers. These outliers will be addressed using capping at the 1st and 99th percentile calculated from the training data only.

## Visualize Outliers Before Capping

```{r visualize-outliers}
# Insulin
ggplot(train_data, aes(x = factor(Diagnosis), y = Insulin,
                       fill = factor(Diagnosis))) +
  geom_boxplot(outlier.color = "red", outlier.size = 1.5, alpha = 0.7) +
  scale_x_discrete(labels = c("0" = "No GDM", "1" = "GDM")) +
  scale_fill_manual(values = c("steelblue", "tomato"),
                    name   = "Diagnosis") +
  labs(title    = "Insulin Outliers Before Capping",
       subtitle = "Red dots are flagged by the IQR rule",
       x = "Diagnosis", y = "Insulin") +
  theme_minimal()

# SkinThickness
ggplot(train_data, aes(x = factor(Diagnosis), y = SkinThickness,
                       fill = factor(Diagnosis))) +
  geom_boxplot(outlier.color = "red", outlier.size = 1.5, alpha = 0.7) +
  scale_x_discrete(labels = c("0" = "No GDM", "1" = "GDM")) +
  scale_fill_manual(values = c("steelblue", "tomato"),
                    name   = "Diagnosis") +
  labs(title    = "SkinThickness Outliers Before Capping",
       subtitle = "Red dots are flagged by the IQR rule",
       x = "Diagnosis", y = "SkinThickness") +
  theme_minimal()

# Pedigree
ggplot(train_data, aes(x = factor(Diagnosis), y = Pedigree,
                       fill = factor(Diagnosis))) +
  geom_boxplot(outlier.color = "red", outlier.size = 1.5, alpha = 0.7) +
  scale_x_discrete(labels = c("0" = "No GDM", "1" = "GDM")) +
  scale_fill_manual(values = c("steelblue", "tomato"),
                    name   = "Diagnosis") +
  labs(title    = "Pedigree Outliers Before Capping",
       subtitle = "Red dots are flagged by the IQR rule",
       x = "Diagnosis", y = "Pedigree") +
  theme_minimal()

# Age
ggplot(train_data, aes(x = factor(Diagnosis), y = Age,
                       fill = factor(Diagnosis))) +
  geom_boxplot(outlier.color = "red", outlier.size = 1.5, alpha = 0.7) +
  scale_x_discrete(labels = c("0" = "No GDM", "1" = "GDM")) +
  scale_fill_manual(values = c("steelblue", "tomato"),
                    name   = "Diagnosis") +
  labs(title    = "Age Outliers Before Capping",
       subtitle = "Red dots are flagged by the IQR rule",
       x = "Diagnosis", y = "Age") +
  theme_minimal()

# BloodPressure
ggplot(train_data, aes(x = factor(Diagnosis), y = BloodPressure,
                       fill = factor(Diagnosis))) +
  geom_boxplot(outlier.color = "red", outlier.size = 1.5, alpha = 0.7) +
  scale_x_discrete(labels = c("0" = "No GDM", "1" = "GDM")) +
  scale_fill_manual(values = c("steelblue", "tomato"),
                    name   = "Diagnosis") +
  labs(title    = "BloodPressure Outliers Before Capping",
       subtitle = "Red dots are flagged by the IQR rule",
       x = "Diagnosis", y = "BloodPressure") +
  theme_minimal()

# BMI
ggplot(train_data, aes(x = factor(Diagnosis), y = BMI,
                       fill = factor(Diagnosis))) +
  geom_boxplot(outlier.color = "red", outlier.size = 1.5, alpha = 0.7) +
  scale_x_discrete(labels = c("0" = "No GDM", "1" = "GDM")) +
  scale_fill_manual(values = c("steelblue", "tomato"),
                    name   = "Diagnosis") +
  labs(title    = "BMI Outliers Before Capping",
       subtitle = "Red dots are flagged by the IQR rule",
       x = "Diagnosis", y = "BMI") +
  theme_minimal()

# Pregnancies
ggplot(train_data, aes(x = factor(Diagnosis), y = Pregnancies,
                       fill = factor(Diagnosis))) +
  geom_boxplot(outlier.color = "red", outlier.size = 1.5, alpha = 0.7) +
  scale_x_discrete(labels = c("0" = "No GDM", "1" = "GDM")) +
  scale_fill_manual(values = c("steelblue", "tomato"),
                    name   = "Diagnosis") +
  labs(title    = "Pregnancies Outliers Before Capping",
       subtitle = "Red dots are flagged by the IQR rule",
       x = "Diagnosis", y = "Pregnancies") +
  theme_minimal()
```

## Cap Outliers Using 1st and 99th Percentile
```{r capping}
numeric_vars <- names(train_data)[names(train_data) != "Diagnosis"]

# Store cap boundaries from training data
cap_lower <- setNames(
  sapply(numeric_vars, function(v) quantile(train_data[[v]], 0.01)),
  numeric_vars
)
cap_upper <- setNames(
  sapply(numeric_vars, function(v) quantile(train_data[[v]], 0.99)),
  numeric_vars
)

# Apply capping to training data
for (var in numeric_vars) {
  train_data[[var]] <- pmax(pmin(train_data[[var]], cap_upper[[var]]), cap_lower[[var]])
}

# Apply same boundaries to test data
for (var in numeric_vars) {
  test_data[[var]] <- pmax(pmin(test_data[[var]], cap_upper[[var]]), cap_lower[[var]])
}

# Confirm no missing values
cat("Missing in training after capping:\n")
print(colSums(is.na(train_data)))

# Confirm zero outliers using same boundaries
cat("\nOutlier counts in training after capping:\n")
print(setNames(sapply(numeric_vars, function(v) {
  sum(train_data[[v]] < cap_lower[[v]] | train_data[[v]] > cap_upper[[v]])
}), numeric_vars))

cat("\nOutlier counts in test after capping:\n")
print(setNames(sapply(numeric_vars, function(v) {
  sum(test_data[[v]] < cap_lower[[v]] | test_data[[v]] > cap_upper[[v]])
}), numeric_vars))
```
##Interpretation-
Before capping, the boxplots showed several outliers flagged by the IQR test, with insulin having the most extreme spread. To reduce the influence of the extreme values while still preserving these records, the numeric predictors were capped at the first and 99th percentiles. These boundaries were calculated on the training data only and then applied to both the training and testing sets. With this approach, all 537 training observations and 231 test observations are preserved since capping adjusts extreme values rather than removing rows.


#Step 3: Feature Selection and Data Processing
# Correlation of each predictor with Diagnosis
```{r}
# 3a: Correlation Analysis

numeric_data  <- train_data[sapply(train_data, is.numeric)]
cor_matrix    <- cor(numeric_data, use = "pairwise.complete.obs")

cor_with_diag <- data.frame(
  Variable    = rownames(cor_matrix),
  Correlation = round(cor_matrix[, "Diagnosis"], 3)
)

# Remove Diagnosis from the list
cor_with_diag <- cor_with_diag[cor_with_diag$Variable != "Diagnosis", ]

# Sort by absolute correlation
cor_with_diag <- cor_with_diag[order(abs(cor_with_diag$Correlation),
                                      decreasing = TRUE), ]
print(cor_with_diag)

```
##Interpretation-
Correlation analysis was used to explore the relationship between each predictor and GDM diagnosis. Glucose showed the strongest association at 0.504, suggesting it is the most linearly related predictor in this sample, consistent with the broader literature [4]. BMI followed at 0.335, Age at 0.232, and SkinThickness and Insulin at 0.225 and 0.221 respectively, though the latter two are likely underestimates given their heavy missingness and median imputation. Pregnancies, BloodPressure, and Pedigree showed the weakest associations ranging from 0.157 to 0.198. All eight predictors showed positive relationship  with Diagnosis. It is worth noting that correlation is a limited measure with a binary outcome as it only captures individual linear relationships and does not reflect how predictors interact within a model.

```{r}
# Full logistic regression model with all predictors
full_model <- glm(
  Diagnosis ~ Pregnancies + Glucose + BloodPressure + SkinThickness +
    Insulin + BMI + Pedigree + Age,
  data   = train_data,
  family = binomial
)

# Stepwise selection using AIC in both directions
# Starts with full model and removes or adds variables
# based on which combination gives the lowest AIC
stepwise_model <- step(full_model, direction = "both", trace = 0)

# Show the final selected model
summary(stepwise_model)
```
##Interpretation-
Stepwise logistic regression selected four predictors from the original eight. BloodPressure, SkinThickness, Insulin, and Age were removed because they did not meaningfully improve model fit based on AIC. The final model retained Glucose, BMI, Pregnancies, and Pedigree, all showing statistically significant positive coefficients. Glucose showed the strongest evidence with a z value of 8.768, followed by BMI at 5.243 and Pregnancies at 3.807. Pedigree was retained with moderate significance. The final AIC of 508.15 confirms this combination of four predictors provides the most favourable balance between model fit and complexity for our specific sample. It is not a guarantee that in a different data sets, stepwsie would select the same features. 

The four variables selected by stepwise are well supported in the literature. Glucose, BMI, and Pregnancies are consistently identified as strong GDM predictors across multiple studies [4][9]. However the dataset has notable limitations. Insulin and SkinThickness had nearly 49% and 30% missing data respectively, which likely weakened their predictive signal and contributed to their removal by stepwise. The dataset also lacks demographic variables such as ethnicity and socioeconomic status which Kokori et al. [10] identify as important for building population specific models. Additionally the Pedigree variable is an indirect measure of family history rather than a direct clinical indicator, which may limit its interpretability in a clinical setting

```{r}
# 3b: Standardization
# Apply z-score normalization to the four selected variables
# Calculate mean and SD from training data only
# Apply same values to test data to prevent leakage

selected_vars <- c("Pregnancies", "Glucose", "BMI", "Pedigree")

train_means <- sapply(train_data[selected_vars], mean)
train_sds   <- sapply(train_data[selected_vars], sd)

# Scale training data
train_data[selected_vars] <- scale(train_data[selected_vars])

# Scale test data using training mean and SD
test_data[selected_vars] <- scale(test_data[selected_vars],
                                   center = train_means,
                                   scale  = train_sds)

# Confirm scaling worked
round(colMeans(train_data[selected_vars]), 3)
round(apply(train_data[selected_vars], 2, sd), 3)

```
##Interpretation-
Z-score normalization was applied to the four variables selected by stepwise: Glucose, BMI, Pregnancies, and Pedigree. The mean and standard deviation were calculated from the training data only and then applied to both the training and test sets. This ensures the test data never influences the scaling parameters, consistent with the leakage prevention approach recommended by Netayawijit et al. [16]. After scaling all four variables have a mean of zero and a standard deviation of one, placing them on a comparable scale regardless of their original units. This step is important because the selected predictors use very different scales. For example Glucose ranges in the hundreds while Pedigree ranges between 0 and 2. Without standardization variables with larger numeric ranges could dominate the model and produce unreliable coefficient estimates.
Furthermore, standarization is applied to make coefficients and odds ratios more comparable across features, showing the effect as one standard deviation increase in each feature. 

```{r}
# 3c: Address Class Imbalance
# Oversample the minority GDM class in training data only
# Test set stays untouched to reflect real world conditions

set.seed(123)

gdm_rows    <- train_data[train_data$Diagnosis == 1, ]
no_gdm_rows <- train_data[train_data$Diagnosis == 0, ]

cat("Before oversampling:\n")
table(train_data$Diagnosis)

# Sample GDM rows with replacement to match No GDM count
gdm_upsampled <- gdm_rows[sample(nrow(gdm_rows),
                                  size    = nrow(no_gdm_rows),
                                  replace = TRUE), ]

# Combine balanced training set
train_balanced <- rbind(no_gdm_rows, gdm_upsampled)

cat("\nAfter oversampling:\n")
table(train_balanced$Diagnosis)

```
##Interpretation-
Oversampling was applied to the training set to address the 65/35 class imbalance identified in step 2c. The minority GDM class was upsampled by randomly duplicating existing GDM rows with replacement until both classes had equal representation of 350 observations each. This gives the model an equal opportunity to learn patterns from both GDM and No GDM patients during training. The test set was left untouched at its original 150 No GDM and 81 GDM distribution to reflect real world conditions and provide an honest evaluation of model performance. This approach is consistent with Du et al. [15] who highlighted class imbalance as a common challenge in GDM datasets and recommended balancing techniques to prevent models from defaulting to the majority class.


```{r}
# Step 4: Model Building and Validation

# 4a: Data Partitioning
# The 70/30 train/test split was completed in Step 2d
# train_balanced contains 700 observations (350 No GDM, 350 GDM)
# test_data contains 231 observations (150 No GDM, 81 GDM)
# train_balanced is used for model training
# test_data is locked until final evaluation

# Convert Diagnosis to factor for classification
train_balanced$Diagnosis <- factor(train_balanced$Diagnosis,
                                    levels = c(0, 1),
                                    labels = c("No_GDM", "GDM"))

test_data$Diagnosis <- factor(test_data$Diagnosis,
                               levels = c(0, 1),
                               labels = c("No_GDM", "GDM"))

# Confirm class distributions
cat("Training class distribution:\n")
table(train_balanced$Diagnosis)

cat("\nTest class distribution:\n")
table(test_data$Diagnosis)

```
##Interpretation-
Data partitioning was completed in Step 2d using a 70/30 split with a fixed seed of 123 for reproducibility. The training set contains 537 observations and the test set contains 231 observations. A 70/30 split was chosen over 80/20 because the small dataset of 768 observations would leave too few positive GDM cases in the test set for reliable evaluation. Oversampling was applied to the training set only to address the class imbalance, resulting in a balanced training set of 700 observations with 350 No GDM and 350 GDM cases. The test set was left at its original distribution of 150 No GDM and 81 GDM to reflect real world conditions.


```{r}
# 4b: Model Development
# Model 1: Logistic Regression
# Using the four variables selected by stepwise in Step 3

logistic_model <- glm(
  Diagnosis ~ Pregnancies + Glucose + BMI + Pedigree,
  data   = train_balanced,
  family = binomial
)

summary(logistic_model)

# Predict on test data
logistic_probs <- predict(logistic_model,
                           newdata = test_data,
                           type    = "response")

# Convert probabilities to class predictions using 0.5 threshold
logistic_preds <- ifelse(logistic_probs > 0.5, "GDM", "No_GDM")
logistic_preds <- factor(logistic_preds, levels = c("No_GDM", "GDM"))

# Confusion matrix
cat("Logistic Regression Confusion Matrix:\n")
table(Predicted = logistic_preds, Actual = test_data$Diagnosis)

```
##Interpretation-
A logistic regression model was trained on the four stepwise selected predictors: Glucose, BMI, Pregnancies, and Pedigree. All four showed statistically significant positive coefficients confirming that higher values increase GDM risk. Glucose was the strongest predictor with a coefficient of 1.254 consistent with Zhang et al. [4]. On the test set the model correctly classified 118 No GDM and 54 GDM patients with 32 false positives and 27 false negatives. The 27 missed GDM cases are the most clinically concerning since missing positive cases in a screening setting could delay important interventions. Overall accuracy was 74.5% with recall of 66.7% and precision of 62.8%. These results will be compared against the random forest model next.

The logistic regression model shows Pregnancy, Glucose, BMI and Pedigree are all positive associated with GDM diagnosis.(See coefficients/odds ratio interpretation below) 



```{r}
# Additional analysis: testing alternative thresholds
# This does not change the official 0.5 threshold result above
# It only explores how recall and precision shift at other cutoffs

test_threshold <- function(threshold) {
  preds <- ifelse(logistic_probs > threshold, "GDM", "No_GDM")
  preds <- factor(preds, levels = c("No_GDM", "GDM"))

  TP <- sum(preds == "GDM"    & test_data$Diagnosis == "GDM")
  TN <- sum(preds == "No_GDM" & test_data$Diagnosis == "No_GDM")
  FP <- sum(preds == "GDM"    & test_data$Diagnosis == "No_GDM")
  FN <- sum(preds == "No_GDM" & test_data$Diagnosis == "GDM")

  accuracy  <- (TP + TN) / (TP + TN + FP + FN)
  precision <- TP / (TP + FP)
  recall    <- TP / (TP + FN)

  cat("Threshold:", threshold, "\n")
  cat("Accuracy: ",  round(accuracy  * 100, 1), "%\n")
  cat("Precision:",  round(precision * 100, 1), "%\n")
  cat("Recall:   ",  round(recall    * 100, 1), "%\n\n")
}

test_threshold(0.5)
test_threshold(0.4)
test_threshold(0.3)
```
##Interpretation-
Three classification thresholds were tested to examine how precision and recall shift as the decision boundary changes.

At the default threshold of 0.5 the model achieved 74.5% accuracy, 62.8% precision, and 66.7% recall. Lowering the threshold to 0.4 kept accuracy at the same 74.5% while recall jumped from 66.7% to 84.0%, with only a small drop in precision from 62.8% to 59.6%. Going further down to 0.3 pushed recall marginally higher to 85.2% but accuracy fell to 68.0% and precision dropped to 52.7%. 

In a clinical screening setting missing a GDM case is far more costly than flagging a patient who turns out not to have it, so the 0.4 threshold stands out as the most practical choice. It catches 17 additional true GDM cases compared to 0.5 with no accuracy loss and only a minor precision decrease. 

The 0.5 threshold is kept as the official result for model comparison purposes since it allows fair comparison against published benchmarks, but a threshold of 0.4 is recommended when deploying this model in a real clinical environment.

```{r}
# Model 2: Random Forest
library(randomForest)

set.seed(123)

rf_model <- randomForest(
  Diagnosis ~ Pregnancies + Glucose + BMI + Pedigree,
  data       = train_balanced,
  ntree      = 500,
  mtry = 3,
  importance = TRUE
)

print(rf_model)

# Predict on test data
rf_preds <- predict(rf_model, newdata = test_data)

# Confusion matrix
cat("Random Forest Confusion Matrix:\n")
table(Predicted = rf_preds, Actual = test_data$Diagnosis)

# Variable importance
cat("\nVariable Importance:\n")
importance(rf_model)
varImpPlot(rf_model)

```
To keep features constant, both logistic regression and random forest are trained using the same 4 variables. However, we understand that this may disadvantage the full potential of random forest, as the it might find other variables useful based on its capabilities for non-linear relationships. 

These results, show an OOB error rate of 11.14% and that the random forest model is better at identifying GDM cases than non-GDM, however these is based on the oversampled trainig data, and therefore it's taken with caution. For Random Forest, Glucose was the most important predictor, followed by BMI.
```{r}
# 4c: Model Evaluation
library(pROC)

# Logistic Regression Performance Metrics
TP <- 54
TN <- 118
FP <- 32
FN <- 27

accuracy  <- (TP + TN) / (TP + TN + FP + FN)
precision <- TP / (TP + FP)
recall    <- TP / (TP + FN)
f1        <- 2 * (precision * recall) / (precision + recall)
f2        <- 5 * (precision * recall) / (4 * precision + recall)
specificity <- TN / (TN + FP)

cat("Logistic Regression Performance Metrics:\n")
cat("Accuracy: ", round(accuracy  * 100, 1), "%\n")
cat("Precision:", round(precision * 100, 1), "%\n")
cat("Recall:   ", round(recall    * 100, 1), "%\n")
cat("Specificity:", round(specificity *100, 1),"%\n")
cat("F1 Score: ", round(f1        * 100, 1), "%\n")
cat("F2 Score: ", round(f2        * 100, 1), "%\n")

# Random Forest Performance Metrics
TP <- 42
TN <- 126
FP <- 24
FN <- 39

accuracy  <- (TP + TN) / (TP + TN + FP + FN)
precision <- TP / (TP + FP)
recall    <- TP / (TP + FN)
f1        <- 2 * (precision * recall) / (precision + recall)
f2        <- 5 * (precision * recall) / (4 * precision + recall)
specificity <- TN / (TN + FP)

cat("\nRandom Forest Performance Metrics:\n")
cat("Accuracy: ", round(accuracy  * 100, 1), "%\n")
cat("Precision:", round(precision * 100, 1), "%\n")
cat("Recall:   ", round(recall    * 100, 1), "%\n")
cat("Specificity:", round(specificity *100, 1),"%\n")
cat("F1 Score: ", round(f1        * 100, 1), "%\n")
cat("F2 Score: ", round(f2        * 100, 1), "%\n")

# ROC-AUC for both models
logistic_roc <- roc(test_data$Diagnosis,
                    logistic_probs,
                    levels = c("No_GDM", "GDM"))

rf_probs <- predict(rf_model,
                    newdata = test_data,
                    type    = "prob")[, "GDM"]

rf_roc <- roc(test_data$Diagnosis,
              rf_probs,
              levels = c("No_GDM", "GDM"))

cat("\nLogistic Regression AUC:", round(auc(logistic_roc), 3), "\n")
cat("Random Forest AUC:      ", round(auc(rf_roc), 3), "\n")

# Plot both ROC curves together
plot(logistic_roc, col = "steelblue", lwd = 2,
     main = "ROC Curves: Logistic Regression vs Random Forest")
plot(rf_roc, col = "tomato", lwd = 2, add = TRUE)
legend("bottomright",
       legend = c(paste("Logistic Regression AUC =", round(auc(logistic_roc), 3)),
                  paste("Random Forest AUC =",       round(auc(rf_roc), 3))),
       col    = c("steelblue", "tomato"),
       lwd    = 2)

```
##Interpretation-
Both models were evaluated on the test set of 231 observations. Logistic regression outperformed random forest on accuracy (74.5% vs 72.7%), recall (66.7% vs 51.9%), F1 (64.7% vs 57.1%), F2 (65.9% vs 53.8%), and AUC (0.837 vs 0.797), while precision was nearly identical for both at around 63%.Recall is the most clinically important metric here since missing a GDM case means a patient misses timely care. Logistic regression caught 54 of 81 actual GDM cases compared to only 42 for random forest. 

The higher AUC for logistic regression confirms it has better overall discriminative ability across all classification thresholds, not just at the default 0.5 cutoff.The AUC score is comparable to what Zhang et al.[4] found at 0.849. When looking at recall, 66.4% while [4] reported 0.69% and specificity at 78% where [4] obtained 75%.

This result differs from broader literature trends where ensemble methods often outperform logistic regression on larger datasets [4][5][6]. With only 768 total observations, random forest likely lacked enough data to learn stable splits and may have overfit, while logistic regression's simpler structure generalized better to this small sample. This aligns with Zhang et al. [4], who note logistic regression remains a strong baseline when sample size limits the advantage of more complex models.


```{r}
# 4d: Model Calibration and Interpretation

# Odds Ratios from Logistic Regression
# Exponentiating the coefficients converts log-odds to odds ratios

odds_ratios <- exp(coef(logistic_model))
conf_int    <- exp(confint(logistic_model))

odds_table <- data.frame(
  Variable    = names(odds_ratios),
  Odds_Ratio  = round(odds_ratios, 3),
  Lower_CI    = round(conf_int[, 1], 3),
  Upper_CI    = round(conf_int[, 2], 3)
)

print(odds_table)
```
```{r}
odds_pregnancy <- exp(0.47092)
odds_glucose <- exp(1.25423)
odds_bmi <- exp(0.61824)
odds_pedigree <- exp(0.22164)

cat("Each 1 SD increase in pregnancy multiplies the odds of GDM by", odds_pregnancy, "\n")
cat("Each 1 SD increase in glucose multiplies the odds of GDM by", odds_glucose, "\n")
cat("Each 1 SD increase in BMI multiplies the odds of GDM by", odds_bmi, "\n")
cat("Each 1 SD increase in pedigree multiplies the odds of GDM by", odds_pedigree, "\n")
```
##Interpretation-
Odds ratios were calculated by exponentiating the logistic regression coefficients to translate log-odds into a more clinically interpretable scale. All four predictors showed odds ratios greater than 1 with confidence intervals that do not cross 1, confirming each is a statistically significant risk factor for GDM. Glucose had by far the strongest effect with an odds ratio of 3.505, meaning each one standard deviation increase in glucose more than triples the odds of a GDM diagnosis. This aligns with Zhang et al. [4] who identified glucose as the dominant clinical predictor across GDM studies. BMI followed with an odds ratio of 1.856 and Pregnancies at 1.601, both consistent with established GDM risk factors in the literature [4][9]. Pedigree showed the weakest but still significant effect at 1.248, reflecting a modest contribution of family history to GDM risk in this dataset.

```{r}
# Calibration Plot
library(ggplot2)

calibration_df <- data.frame(
  predicted = logistic_probs,
  actual    = as.numeric(test_data$Diagnosis == "GDM")
)

# Bin predictions into deciles
calibration_df$bin <- cut(calibration_df$predicted,
                          breaks = seq(0, 1, by = 0.1),
                          include.lowest = TRUE)

calibration_summary <- aggregate(
  cbind(predicted, actual) ~ bin,
  data = calibration_df,
  FUN  = mean
)

print(calibration_summary)

# Plot calibration curve
ggplot(calibration_summary, aes(x = predicted, y = actual)) +
  geom_point(size = 3, color = "steelblue") +
  geom_line(color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Calibration Plot: Logistic Regression",
       subtitle = "Dashed line represents perfect calibration",
       x = "Mean Predicted Probability",
       y = "Observed Proportion of GDM Cases") +
  theme_minimal()

```
```{r brier-score}
#Brier score for logistic regression probability predictions
actual_gdm <- ifelse(as.character(test_data$Diagnosis) %in% c("1", "GDM"), 1, 0)

brier_score <- mean((logistic_probs - actual_gdm)^2)

cat("Logistic Regression Brier Score:", round(brier_score, 3), "\n")
```
##Interpretation-
The calibration analysis grouped predicted probabilities into ten bins and compared the mean predicted probability in each bin to the observed proportion of actual GDM cases. The model showed strong calibration at the extremes, with the lowest and highest probability bins closely matching their observed outcomes. However several middle bins, particularly the 0.3 to 0.4 and 0.6 to 0.7 ranges, showed noticeable overconfidence where predicted probabilities exceeded the observed proportion of positive cases. This is likely a result of the relatively small test set of 231 observations, which leaves few patients per bin and makes the observed proportions less stable. Despite this variability in the middle ranges, the model demonstrates reasonable overall calibration, particularly in the high confidence ranges most relevant to clinical decision making. The added Brier score provides a numeric view of the model's probability prediction error, with a lower score indicating better prediction performance [21]. With a score of 0.162, the model demonstrates reasonable calibration despite some variability in the middle ranges. 


```{r}
# 4e: Multiple Testing Correction

# Extract p-values from the logistic regression model
# excluding the intercept
p_values <- summary(logistic_model)$coefficients[-1, 4]

# Apply Bonferroni correction
bonferroni_adjusted <- p.adjust(p_values, method = "bonferroni")

# Apply Benjamini-Hochberg (FDR) correction as a less conservative alternative
fdr_adjusted <- p.adjust(p_values, method = "BH")

correction_table <- data.frame(
  Variable          = names(p_values),
  Raw_p_value       = round(p_values, 5),
  Bonferroni        = round(bonferroni_adjusted, 5),
  FDR_BH            = round(fdr_adjusted, 5)
)

print(correction_table)

```
##Interpretation-
Multiple testing correction was applied to the four predictor p-values from the logistic regression model using both Bonferroni and Benjamini-Hochberg (FDR) methods. Pregnancies, Glucose, and BMI remained highly significant under both corrections with adjusted p-values near zero. Pedigree showed a raw p-value of 0.0297 which is significant at the conventional 0.05 threshold, but after the more conservative Bonferroni correction the adjusted p-value rose to 0.11881, no longer reaching significance. Under the FDR correction, however, Pedigree remained significant at 0.0297. Given that Bonferroni can be overly conservative when testing a small number of correlated clinical variables, the FDR result is considered more appropriate for this analysis, supporting the retention of Pedigree as a meaningful, if modest, predictor of GDM risk consistent with its established role as a family history indicator in the literature [4][9].

However, since the main purpose of the mdoel is prediction, test-set performance in recall/sensitivity and calibration are more important than p-values alone. Furthermore, these p-values are taken from the selected stepwise features they are not full evidence and p-values might look more stable than they are. 


**Step 5: Analysis, Limitations, and Reporting**

**5a: Model Limitations**

*Strengths*:
The logistic regression model demonstrated solid performance for a small clinical dataset, achieving an AUC of 0.837 which exceeds the pooled AUROC of 0.79 reported for logistic regression models across 25 GDM studies by Zhang et al. [4]. The model is interpretable, with clinically meaningful odds ratios that align with established GDM risk factors in the literature. The preprocessing pipeline followed best practices including leakage-free imputation, capping, standardization from training data only, and oversampling strictly on the training set. The four selected predictors, Glucose, BMI, Pregnancies, and Pedigree, are all well supported by the literature as genuine GDM risk factors [4][9].

*Limitations*:
The most significant limitation is dataset size. With only 768 observations and 268 positive GDM cases, the model was trained on a relatively small sample which limits its statistical power and reliability. Yitayeh et al. [14] specifically flag small sample size as one of the most common methodological weaknesses in published GDM prediction studies, noting that machine learning models generally require larger datasets to reach their full potential.

The heavy missingness in Insulin at 48.7% and SkinThickness at 29.6% is a second major concern. Both variables were imputed using median values, which reduces their natural variability and likely weakened their predictive signal. This contributed to both being dropped by stepwise selection, even though they carry clinical relevance as metabolic indicators.It is also important to mention the limitations brought by the unbalanced data and utilizing oversampling, where duplicating existing observations can increase overfitting and distort the results from probability calibration. 

The dataset also lacks demographic variables. There is no information on ethnicity, socioeconomic status, or healthcare access. Kokori et al. [10] and Chen et al. [17] both emphasize that the absence of demographic context makes it impossible to audit whether the model performs differently across patient subgroups. A model that appears accurate overall may still produce disparate outcomes for specific populations, which is a serious concern in a clinical setting.

Finally the dataset is known to originate from a study of Pima Indian women, as identified by Chen et al. [17]. This means the model was trained on a single homogeneous population and cannot be assumed to generalize to women from different ethnic, geographic, or socioeconomic backgrounds. Kokori et al. [10] specifically warn against applying GDM prediction models trained on one population to groups that are ethnically or demographically different without external validation.

Given all these limitations and at a sensitivity of 66.7% and identifying only 54 of 81 true GDM cases, in a field where false negatives are costly,our logistic regression model can not be used or framed as a full diagnostic model, but instead should be used as a decision-support tool. Before clinical use, cross validation, external validation, threshold tuning, and the other recommendation explained below should be considered. 

*Generalizability*:
The model should be considered preliminary. Zhang et al. [4] found that only 4 of 25 reviewed GDM studies tested their models on external data, which means most published models including this one have not been validated beyond their training population. Before this model could be considered for clinical use it would need to be tested on an external dataset drawn from a different population and setting to confirm that its performance holds up in the real world.

*5b: Recommendations for Improvement*

1. Larger and more diverse dataset

The most impactful improvement would be collecting or obtaining a larger dataset with greater demographic diversity. Studies like Artzi et al. used over 580,000 pregnancies from electronic health records and demonstrated far more reliable and generalizable models. A larger dataset would also reduce the need for heavy imputation and allow more complex models like random forest and gradient boosting to perform closer to their potential.

2. Better imputation for high-missingness variables

Median imputation for Insulin and SkinThickness was a practical choice but not ideal. MICE with predictive mean matching as recommended by Netayawijit et al. [16] would produce more realistic imputed values by using the relationships between variables rather than a single summary statistic. This could restore some of the predictive signal lost during imputation.

3. Include demographic variables

Adding variables like ethnicity, family history of diabetes, socioeconomic status, and prior GDM history would bring the dataset closer to what the literature identifies as the most comprehensive predictor set [4][9]. It would also enable a fairness audit across subgroups as recommended by Chen et al. [17].

4. Try additional models

Gradient boosting and XGBoost were identified by Ye et al. [5] and Watanabe et al. [6] as consistently strong performers in GDM prediction. With a larger dataset these models would likely outperform logistic regression. Adding them to the comparison would give a more complete picture of model options.

5. Explainability with SHAP

Du et al. [15] demonstrated that SHAP values significantly improve the clinical trustworthiness of GDM prediction models by providing both global and individual level explanations. Adding SHAP to the logistic regression or any future model would help clinicians understand why a specific patient was flagged as high risk, which is essential for adoption in a real clinical environment.

6. Cross validation

The current evaluation relied on a single 70/30 train/test split. Using k-fold cross validation during training would produce more stable and reliable performance estimates by averaging results across multiple splits rather than depending on one particular random partition.


**Step 6: Ethical Implications**

Developing a predictive model for gestational diabetes in a clinical setting carries ethical responsibilities that extend beyond model accuracy. A model that influences clinical decisions about pregnant women must be held to a higher standard than a typical data science project, and several ethical considerations are directly relevant to this analysis.

*Algorithmic Fairness and Bias*

The most immediate ethical concern is whether the model performs equally well across different patient populations. Chen et al. [17] provide a detailed framework for thinking about algorithmic fairness in healthcare AI and note that assessments of models across patient subgroups have revealed meaningful inequalities in how patients are diagnosed and treated. This dataset originates from a study of Pima Indian women, a single homogeneous population, and contains no demographic variables such as ethnicity, socioeconomic status, or healthcare access. This makes it impossible to audit whether the model produces disparate outcomes for women from different backgrounds. A model that appears accurate overall may still systematically miss GDM cases in certain subgroups, which would cause real harm in a clinical setting.

Kokori et al. [10] make the same point specifically in the context of GDM research, noting that applying predictive guidelines developed in one population to groups that are ethnically and socioeconomically different tends to produce unequal outcomes. Before this model could be considered for deployment, fairness audits across demographic subgroups would be essential.

*Clinical Decision Support vs Automated Diagnosis*

A model like this should never be positioned as a replacement for clinical judgment. Du et al. [15] are explicit about this in their work on explainable GDM prediction, framing the entire purpose of the system around supporting clinicians rather than replacing them. The model flags elevated risk for further clinical attention. The final decision about diagnosis, monitoring, and intervention must remain with the healthcare provider. Deploying a model that clinicians cannot question or override would be ethically indefensible in a maternal health context.

*Transparency and Explainability*
Patients have a right to understand why they are being flagged as high risk. A black box model that produces a risk score without explanation does not meet the transparency standard expected in clinical care. Du et al. [15] applied SHAP values to their GDM prediction system specifically to increase trustworthiness and acceptability to clinicians, noting that models that cannot explain individual predictions are unlikely to be adopted regardless of their accuracy. Any clinical deployment of a model like this should include explanations of which factors contributed most to an individual patient's risk score.
Consent and Data Privacy
Patients whose data is used to train a predictive model should be informed and their consent obtained. This is not just a legal requirement under frameworks like HIPAA but an ethical obligation. The source and composition of the training data should be documented and disclosed so that clinicians and patients can make informed judgments about how much weight to give the model's predictions.
Consequences of False Negatives
As discussed in section 4c, false negatives carry the highest clinical cost in a GDM screening context. Missing a positive case means a patient does not receive timely dietary guidance, glucose monitoring, or clinical intervention during pregnancy, which can lead to complications for both mother and child. Any deployment decision must account for this asymmetry in error costs by adjusting the classification threshold to prioritize recall, accepting more false positives rather than missing true positives.
