---
title: "Gestational Diabetes Risk Prediction"
author: 'Group 3: Manav, Davis, Daniela, Abenezer'
date: "2026-06-09"
output: word_document
---
##Step 2

# 2a
```{r load/view}
patients <- read.csv("patients.csv")

head(patients)
```

```{r inspect-structure}
str(patients)
```
First, we load the data set into R and inspect the structure of the data set to review the number of observations, variables, variable names, and data types.

```{r summary-stats}
summary(patients)
```
```{r standard-deviation}
sapply(patients, sd)
```
Summary statistics were calculated to examine the spread of each variable. The results showed possible data quality issues because several clinical variables had a minimum of zero, including glucose, blood pressure, skin thickness, insulin, and BMI. These values need to be examined more closely during cleaning steps.

# 2c
```{r loading-ggplot2}
library(ggplot2)
```
```{r}
ggplot(patients, aes(x = factor(Diagnosis))) +
  geom_bar() +
  labs(title = "Count of Diagnosis Groups",
       x = "Diagnosis",
       y = "Count")
```

The bar chart shows that there are more patients without a GEDM diagnosis than with a diagnosis. That means the outcome variable is somewhat imbalanced, which should be considered when evaluating classification model performance.

```{r numeric-features-histogram}
ggplot(patients, aes(x = Glucose)) +
  geom_histogram(bins = 30) +
  labs(title = "Distribution of Glucose",
       x = "Glucose",
       y = "Count")

ggplot(patients, aes(x = BloodPressure)) +
  geom_histogram(bins = 30) +
  labs(title = "Distribution of Blood Pressure",
       x = "Blood Pressure",
       y = "Count")

ggplot(patients, aes(x = BMI)) +
  geom_histogram(bins = 30) +
  labs(title = "Distribution of BMI",
       x = "BMI",
       y = "Count")
```

These histograms show that blood pressure, glucose, and BMI are mostly concentrated around the middle of their ranges. Blood pressure is centered mainly around 60 to 80, glucose around 90 to 130, and BMI around 25 to 40. Each plot also shows a small group of values at or near zero, which stands apart from the rest of the distribution. These zeros will need to be addressed during cleaning.

```{r feature-boxplot-by-diagnosis}
ggplot(patients, aes(x = factor(Diagnosis), y = Pregnancies)) +
  geom_boxplot() +
  labs(title = "Pregnancies by Diagnosis",
       x = "Diagnosis",
       y = "Pregnancies")

ggplot(patients, aes(x = factor(Diagnosis), y = Glucose)) +
  geom_boxplot() +
  labs(title = "Glucose by Diagnosis",
       x = "Diagnosis",
       y = "Glucose")

ggplot(patients, aes(x = factor(Diagnosis), y = BloodPressure)) +
  geom_boxplot() +
  labs(title = "Blood Pressure by Diagnosis",
       x = "Diagnosis",
       y = "Blood Pressure")

ggplot(patients, aes(x = factor(Diagnosis), y = SkinThickness)) +
  geom_boxplot() +
  labs(title = "Skin Thickness by Diagnosis",
       x = "Diagnosis",
       y = "Skin Thickness")

ggplot(patients, aes(x = factor(Diagnosis), y = Insulin)) +
  geom_boxplot() +
  labs(title = "Insulin by Diagnosis",
       x = "Diagnosis",
       y = "Insulin")

ggplot(patients, aes(x = factor(Diagnosis), y = BMI)) +
  geom_boxplot() +
  labs(title = "BMI by Diagnosis",
       x = "Diagnosis",
       y = "BMI")

ggplot(patients, aes(x = factor(Diagnosis), y = Pedigree)) +
  geom_boxplot() +
  labs(title = "Pedigree by Diagnosis",
       x = "Diagnosis",
       y = "Pedigree")

ggplot(patients, aes(x = factor(Diagnosis), y = Age)) +
  geom_boxplot() +
  labs(title = "Age by Diagnosis",
       x = "Diagnosis",
       y = "Age")
```
These box plots compare each predictor between patients without a GDM diagnosis, which is designated as 0, and patients with a GDM diagnosis, which is designated as 1. The diagnosed group generally shows higher values for pregnancies, glucose, BMI, pedigree, and age. Glucose shows one of the clearest separations, with the diagnosed group having noticeably higher median and interquartile range. Blood pressure shows only a small difference between the groups, and most importantly, the data quality issue shown here is that skin thickness and insulin may have values concentrated at zero, which indicate missing values or missing measurements.

```{r age/glucose-diagnosis}
ggplot(patients, aes(x = Age, y = Glucose, color = factor(Diagnosis))) +
  geom_point() +
  labs(title = "Age and Glucose by Diagnosis",
       x = "Age",
       y = "Glucose",
       color = "Diagnosis")
```
This scatterplot shows the relationship between age and glucose, colored with the diagnosis. Diagnosed patients appear more frequently at higher glucose values, while non-diagnosed patients appear to be more concentrated at lower to moderate glucose values. Age does not show as clear of a separation as glucose, because both diagnosis groups appear across much of the age range. The plot also shows a few unusual data points, including glucose values of zero and some unusually high ages, including one around the age of 80, which should be reviewed as possible data quality issues.

```{r glucose/bmi-diagnosis}
ggplot(patients, aes(x = BMI, y = Glucose, color = factor(Diagnosis))) +
  geom_point() +
  labs(title = "BMI and Glucose by Diagnosis",
       x = "BMI",
       y = "Glucose",
       color = "Diagnosis")
```
This scatter plot shows the relationship between BMI and glucose, colored by diagnosis. Diagnosed patients appear more common at higher glucose levels, especially above roughly 125 to 140. BMI also appears somewhat higher among diagnosed patients, but the separation is not as clear. The plot shows possible data quality issues as well, including BMI values at zero and glucose values at zero, which are not clinically realistic and, again, should be reviewed during cleaning.

# 2d
```{r checking-for-na}
colSums(is.na(patients))
```
```{r checking-for-0s}
colSums(patients == 0)
```
No variables contained values coded as NA. However, several variables contained values of zero that are likely invalid or represent missing measurements. Zeros for things like pregnancy and diagnosis are valid because they represent no prior pregnancies and no diagnosis. In contrast, zeros for glucose, blood pressure, skin thickness, insulin, and BMI are not realistic and are highly suspicious, so they should be treated as missing values for cleaning.

```{r data-cleaning}
patients_clean <- patients

invalid_zero_vars <- c("Glucose", "BloodPressure", "SkinThickness", "Insulin", "BMI")

patients_clean[invalid_zero_vars] <- lapply(
  patients_clean[invalid_zero_vars],
  function(x) ifelse(x == 0, NA, x)
)
```
```{r}
colSums(is.na(patients_clean))
```
```{r}
compare_imputation_methods <- function(data, target_variable) {
  
  set.seed(123)
  
  predictor_vars <- c("Pregnancies", "Glucose", "BloodPressure", "BMI", "Pedigree", "Age")
  predictor_vars <- predictor_vars[predictor_vars != target_variable]
  
  observed_rows <- which(
    !is.na(data[[target_variable]]) &
      complete.cases(data[predictor_vars])
  )
  
  hide_rows <- sample(observed_rows, size = round(0.30 * length(observed_rows)))
  train_rows <- setdiff(observed_rows, hide_rows)
  
  true_values <- data[[target_variable]][hide_rows]
  
  median_value <- median(data[[target_variable]][train_rows], na.rm = TRUE)
  median_predictions <- rep(median_value, length(hide_rows))
  
  regression_formula <- as.formula(
    paste(target_variable, "~", paste(predictor_vars, collapse = " + "))
  )
  
  regression_model <- lm(regression_formula, data = data[train_rows, ])
  regression_predictions <- predict(regression_model, newdata = data[hide_rows, ])
  
  data.frame(
    Variable = target_variable,
    Hidden_Values = length(hide_rows),
    Median_MAE = mean(abs(true_values - median_predictions)),
    Regression_MAE = mean(abs(true_values - regression_predictions)),
    Median_RMSE = sqrt(mean((true_values - median_predictions)^2)),
    Regression_RMSE = sqrt(mean((true_values - regression_predictions)^2))
  )
}

rbind(
  compare_imputation_methods(patients_clean, "Glucose"),
  compare_imputation_methods(patients_clean, "BloodPressure"),
  compare_imputation_methods(patients_clean, "BMI"),
  compare_imputation_methods(patients_clean, "SkinThickness"),
  compare_imputation_methods(patients_clean, "Insulin")
)
```
For imputation, median imputation was compared to regression for the variables where invalid zeros had been converted to missing values. This comparison was done by temporarily hiding a portion of the known values, imputing those values using each method, and then comparing the imputed values to the true values. MAE measures the average size of the prediction errors, while RMSE also measures prediction error, but gives more weight to bigger mistakes. For both measures, lower values indicate better performance. Regression imputation had lower MAE and RMSE than median imputation for glucose, blood pressure, BMI, skin thickness, and insulin, so regression is the choice for these variables in the cleaned data set. Skin thickness and insulin were not used as predictors for imputing other variables due to their high missingness. Using them as predictors would limit the model to fewer complete rows, so we chose a simpler, more consistent predictor set.

# 2e
```{r outliers-cleaned}
detect_outliers <- function(x) {
  q1 <- quantile(x, 0.25, na.rm = TRUE)
  q3 <- quantile(x, 0.75, na.rm = TRUE)
  iqr <- q3 - q1
  
  lower <- q1 - 1.5 * iqr
  upper <- q3 + 1.5 * iqr
  
  sum(x < lower | x > upper, na.rm = TRUE)
}

sapply(patients_clean, detect_outliers)
```
After running the outlier calculation, we identified possible outliers in pregnancies, blood pressure, skin thickness, insulin, BMI, pedigree, and age. Glucose and diagnosis did not have any outliers. Most values were retained because they may represent real clinical variation rather than data entry errors. Age was noted as a separate concern because some of the high values may not fit the expected GDM screening population. However, no age values were removed or changed at this stage.

##Step 3

#3a
```{r correlation-analysis}
numeric_data <- patients_clean[sapply(patients_clean, is.numeric)]

cor_matrix <- cor(numeric_data, use = "pairwise.complete.obs")

diagnosis_correlations <- data.frame(
  Variable = rownames(cor_matrix),
  Correlation_With_Diagnosis = cor_matrix[, "Diagnosis"]
)

diagnosis_correlations <- diagnosis_correlations[
  diagnosis_correlations$Variable != "Diagnosis",
]

diagnosis_correlations <- diagnosis_correlations[
  order(abs(diagnosis_correlations$Correlation_With_Diagnosis), decreasing = TRUE),
]

diagnosis_correlations
```
For the beginning of step 3, we examined how each predictor variable was associated with diagnosis. Since diagnosis is coded for 1 as a positive GDM diagnosis and 0 for no diagnosis, a positive Pearson correlation indicates that higher values of a predictor are associated with a greater likelihood of getting a positive diagnosis. A correlation close to 0 would suggest that the variable has little individual linear association with diagnosis, while a negative correlation would indicate that higher values are more associated with not having a diagnosis.

```{r feature-selection}
model_data_all <- patients_clean[
  complete.cases(patients_clean[, c(
    "Pregnancies", "Glucose", "BloodPressure", "SkinThickness",
    "Insulin", "BMI", "Pedigree", "Age", "Diagnosis"
  )]),
  c(
    "Pregnancies", "Glucose", "BloodPressure", "SkinThickness",
    "Insulin", "BMI", "Pedigree", "Age", "Diagnosis"
  )
]


#logistic regression model with all predictors
full_model_all <- glm(
  Diagnosis ~ Pregnancies + Glucose + BloodPressure + SkinThickness +
    Insulin + BMI + Pedigree + Age,
  data = model_data_all,
  family = binomial
)

#stepwise feature selection
stepwise_model_all <- step(full_model_all, direction = "both")

summary(stepwise_model_all)
```

We used a stepwise logistic regression as a model-based feature selection method. The full model included all available predictors, which includes pregnancies, glucose, blood pressure, skin thickness, insulin, BMI, pedigree, and age. Stepwise selection used AIC to remove variables that did not improve the model enough to justify including them. The AIC decreased from 362.02 in the full model to 356.89 in the final selected model.
The final selected model included pregnancies, glucose, BMI, pedigree, and age. Blood pressure, insulin, and skin thickness were removed by the stepwise process. Glucose, BMI, and pedigree showed the clearest evidence as predictors, while age and pregnancies were retained but should be interpreted more cautiously.

```{r standardization}
#dataset using only selected predictors
model_data_selected <- patients_clean[
  complete.cases(patients_clean[, c(
    "Pregnancies", "Glucose", "BMI", "Pedigree", "Age", "Diagnosis"
  )]),
  c("Pregnancies", "Glucose", "BMI", "Pedigree", "Age", "Diagnosis")
]

#standardize predictor variables
model_data_standardized <- model_data_selected

model_data_standardized[, c("Pregnancies", "Glucose", "BMI", "Pedigree", "Age")] <-
  scale(model_data_standardized[, c("Pregnancies", "Glucose", "BMI", "Pedigree", "Age")])

summary(model_data_standardized)
```

After feature selection, we standardized the selected predictor variables, which were pregnancies, glucose, BMI, pedigree, and age. Standardization converts each predictor into a z-score-like value by subtracting the variable's mean and dividing by its standard deviation. This means that a value of zero represents an average value for a given predictor. This step is useful because the predictors are measured on different scales. For example, glucose, BMI, age, and pedigree all use different units and ranges. Standardizing them places predictors on a comparable scale before we build the model. This prepares the dataset for step 4.

##Step 4

#4a
```{r data-partitioning}
set.seed(123)

#separate row indexes for each Diagnosis group
index_0 <- which(model_data_selected$Diagnosis == 0)
index_1 <- which(model_data_selected$Diagnosis == 1)

#80% of each group for the training set
train_0 <- sample(index_0, size = floor(0.80 * length(index_0)))
train_1 <- sample(index_1, size = floor(0.80 * length(index_1)))

#combine the training indexes
train_index <- c(train_0, train_1)

#training and testing datasets
train_data <- model_data_selected[train_index, ]
test_data <- model_data_selected[-train_index, ]

dim(train_data)
dim(test_data)

#check Diagnosis proportions in the full, training, and testing data
prop.table(table(model_data_selected$Diagnosis))
prop.table(table(train_data$Diagnosis))
prop.table(table(test_data$Diagnosis))
```

The modeling dataset was split into training and testing sets using an 80-20 split. Because the outcome variable was imbalanced, with 65% as non-GDM cases and 35% as GDM cases, the split was stratified by diagnosis. This ensured that the training and testing sets had nearly identical proportions of non-diagnosed to diagnosed patients. The final split includes 601 training observations and 151 testing observations. This supports a fair model evaluation because the model is tested on unseen data with a distribution similar to our original dataset. Both datasets contain approximately 64.9% non-GDM cases and 35.1% GDM cases.

#4b
```{r modeling}
#logistic regression model using the training data
logistic_model <- glm(
  Diagnosis ~ .,
  data = train_data,
  family = binomial
)

summary(logistic_model)
```

A logistic regression model was built using the training dataset and our selected features. Logistic regression was appropriate because our dependent variable, diagnosis, is binary, and because the logistic regression provides an interpretable baseline for clinical prediction. In our model, pregnancies, glucose, BMI, and pedigree were statistically significant predictors for GDM diagnosis. Age was not statistically significant in this model. The positive coefficients indicate that higher values of these predictors, generally speaking, are associated with higher predicted risk of GDM. These findings align with our literature review, which identified glucose, BMI, pregnancy history, and family history-related risk as important predictors for GDM [4], [9].

#4c
```{r evaluation}
#predict probabilities on the testing data
test_prob <- predict(
  logistic_model,
  newdata = test_data,
  type = "response"
)

#convert probabilities to predicted classes using 0.50 threshold
test_pred <- ifelse(test_prob >= 0.50, 1, 0)

#make actual and predicted values factors with same levels
actual <- factor(test_data$Diagnosis, levels = c(0, 1))
predicted <- factor(test_pred, levels = c(0, 1))

#confusion matrix
conf_matrix <- table(Actual = actual, Predicted = predicted)
conf_matrix

#extract values from confusion matrix
TN <- conf_matrix["0", "0"]
FP <- conf_matrix["0", "1"]
FN <- conf_matrix["1", "0"]
TP <- conf_matrix["1", "1"]

#calculate evaluation metrics
accuracy <- (TP + TN) / sum(conf_matrix)
precision <- TP / (TP + FP)
recall <- TP / (TP + FN)
specificity <- TN / (TN + FP)
f1_score <- 2 * ((precision * recall) / (precision + recall))
f2_score <- 5 * ((precision * recall) / ((4 * precision) + recall))

#put metrics into a table
metrics <- data.frame(
  Accuracy = accuracy,
  Precision = precision,
  Recall = recall,
  Specificity = specificity,
  F1_Score = f1_score,
  F2_Score = f2_score
)

round(metrics, 3)

#calculate ROC-AUC using predicted probabilities
auc_function <- function(actual, predicted_prob) {
  actual_numeric <- as.numeric(as.character(actual))
  ranks <- rank(predicted_prob)
  n_pos <- sum(actual_numeric == 1)
  n_neg <- sum(actual_numeric == 0)
  
  auc <- (sum(ranks[actual_numeric == 1]) - n_pos * (n_pos + 1) / 2) / 
    (n_pos * n_neg)
  
  return(auc)
}

roc_auc <- auc_function(actual, test_prob)
round(roc_auc, 3)
```

The logistic regression model was evaluated on the testing data set using a classification threshold of 50%. At this threshold, the model achieved accuracy of about 77.5%, meaning it correctly classified 77.5% of patients overall. The precision was about 73.2%, meaning that when the model predicted a patient as having GDM, it was correct 73.2% of the time. However, recall was lower at 56.6%, meaning the model correctly identified only 56.6% of the patients who actually had GDM. Specificity was higher at 88.8%, showing that our model was better at correctly identifying non-GDM patients rather than identifying GDM patients. The ROC AUC was 85.9%. This means that the model had a good overall ability to separate GDM and non-GDM cases across possible thresholds. However, the recall value at the default 50% threshold shows that the model still missed a meaningful number of GDM cases. In a clinical context, this is important because failing to identify high-risk patients could delay monitoring or intervention. Therefore, while the model shows useful predictive ability overall, the threshold of 50% may not be the best one if the goal is to prioritize identifying more patients with GDM.

Because our literature review frames this model as a screening tool rather than a final diagnostic tool, recall is especially important. Our review explains that missing a positive GDM case may delay monitoring or intervention, so making false negatives is a major concern in this clinical context [15]. Therefore, the next step is to compare lower classification thresholds to determine whether recall improves while accuracy, precision, and specificity remain acceptable.

```{r threshhold-comparision}
calculate_metrics <- function(threshold) {
  
  predicted_class <- ifelse(test_prob >= threshold, 1, 0)
  
  actual <- factor(test_data$Diagnosis, levels = c(0, 1))
  predicted <- factor(predicted_class, levels = c(0, 1))
  
  cm <- table(Actual = actual, Predicted = predicted)
  
  TN <- cm["0", "0"]
  FP <- cm["0", "1"]
  FN <- cm["1", "0"]
  TP <- cm["1", "1"]
  
  accuracy <- (TP + TN) / sum(cm)
  precision <- TP / (TP + FP)
  recall <- TP / (TP + FN)
  specificity <- TN / (TN + FP)
  f1_score <- 2 * ((precision * recall) / (precision + recall))
  f2_score <- 5 * ((precision * recall) / ((4 * precision) + recall))
  
  data.frame(
    Threshold = threshold,
    Accuracy = accuracy,
    Precision = precision,
    Recall = recall,
    Specificity = specificity,
    F1_Score = f1_score,
    F2_Score = f2_score
  )
}

threshold_results <- rbind(
  calculate_metrics(0.50),
  calculate_metrics(0.45),
  calculate_metrics(0.40),
  calculate_metrics(0.35),
  calculate_metrics(0.30)
)

round(threshold_results, 3)
```

The threshold comparison showed that lowering the classification threshold improved recall, which is important in this clinical context. The default of a 50% threshold had a recall of 56.6%, meaning the model missed a substantial amount of GDM cases. Lowering the threshold to 35% improved recall to 77.4%, while also producing the highest overall accuracy at 79.5%, while also having the highest F1 score at 72.6%. Although the 30% threshold produced the highest recall at 83%, it also reduced precision and specificity substantially. Therefore, the 35% threshold appears to provide the best balance between identifying more actual GDM cases while still maintaining acceptable overall performance from the model.

When compared with the literature review, our logistic regression model showed reasonable predictive performance, especially with a ROC-AUC of 85.9%. However, a direct comparison to published studies should be viewed with caution because these reviewed studies used different datasets, different validation methods, had different sample sizes, and predictors. Our literature review noted that model performance can vary substantially depending on the dataset and validation approach [4]. In addition to this, our dataset is relatively small, lacks demographic variables, and uses a limited set of routinely collected clinical predictors. These differences might explain why our model shows a useful overall separation but still requires a threshold adjustment to improve recall in a clinical screening context.

#4d
```{r odds-ratios}
odds_ratios <- data.frame(
  Predictor = names(coef(logistic_model)),
  Coefficient = coef(logistic_model),
  Odds_Ratio = exp(coef(logistic_model)),
  Lower_95_CI = exp(confint.default(logistic_model)[, 1]),
  Upper_95_CI = exp(confint.default(logistic_model)[, 2]),
  P_Value = summary(logistic_model)$coefficients[, 4]
)

odds_ratios$Coefficient <- round(odds_ratios$Coefficient, 3)
odds_ratios$Odds_Ratio <- round(odds_ratios$Odds_Ratio, 3)
odds_ratios$Lower_95_CI <- round(odds_ratios$Lower_95_CI, 3)
odds_ratios$Upper_95_CI <- round(odds_ratios$Upper_95_CI, 3)
odds_ratios$P_Value <- round(odds_ratios$P_Value, 4)

odds_ratios
```

The odds ratio results helped us interpret the logistic regression model in practical terms. Pregnancies, glucose, BMI, and pedigree were all positively associated with the GDM diagnosis. Holding the other features consistent, each additional pregnancy was associated with 1.131 times higher odds of GDM. Each one-unit increase in glucose was associated with 1.035 times higher odds of GDM, and each one-unit increase in BMI was associated with 1.091 times higher odds. The pedigree variable had an odds ratio of 1.996, suggesting that higher family history-related diabetes risk was associated with substantially higher odds of GDM. These findings align with the literature review, which identified glucose, BMI, pregnancy history, and family history risk as important predictors for GDM [4], [9].

Age had an odds ratio of 1.004 and was not statistically significant, suggesting that age did not add meaningful independent predictive value after accounting for the other features in this model. Although age is commonly identified in the literature as a GDM risk factor, it was retained because it was selected during feature selection and is supported by prior research.

Another important point is that although pedigree had the largest odds ratio, this reflects a one-unit increase in that variable, which may be a large change. Glucose had the strongest statistical evidence in the model, and remains one of the most important predictors, especially because glucose values vary over a wider range and are directly related to a GDM diagnosis.

```{r calibration}
calibration_data <- data.frame(
  Actual = test_data$Diagnosis,
  Predicted_Probability = test_prob
)

#put predicted probabilities into risk bins
calibration_data$Risk_Group <- cut(
  calibration_data$Predicted_Probability,
  breaks = c(0, 0.20, 0.40, 0.60, 0.80, 1),
  include.lowest = TRUE
)

#compare average predicted risk to observed GDM rate in each group
calibration_table <- aggregate(
  cbind(Predicted_Probability, Actual) ~ Risk_Group,
  data = calibration_data,
  FUN = mean
)

group_counts <- aggregate(
  Actual ~ Risk_Group,
  data = calibration_data,
  FUN = length
)

calibration_table$N <- group_counts$Actual

#rename columns
names(calibration_table) <- c(
  "Risk_Group",
  "Average_Predicted_Risk",
  "Observed_GDM_Rate",
  "N"
)
calibration_table$Average_Predicted_Risk <- round(calibration_table$Average_Predicted_Risk, 3)
calibration_table$Observed_GDM_Rate <- round(calibration_table$Observed_GDM_Rate, 3)

calibration_table
```

Calibration was assessed by grouping patients into predicted risk ranges and comparing the average predicted probability with the observed GDM rate in each group. Overall, the model appeared reasonably well-calibrated. In the 20-40%, 40-60%, and 60-80% groups, the average predicted risks were close to the observed GDM rate. The lowest risk group showed some overestimation, while the highest risk group showed underestimation because all patients in that group had GDM. However, it should be noted that the highest risk group only contained 14 patients. Overall, calibration results suggest that the model's predicted probabilities were reasonably consistent with the observed GDM rate.

```{r correction}
#predictor p-values
raw_p_values <- summary(logistic_model)$coefficients[-1, 4]

#Benjamini-Hochberg false discovery rate correction
adjusted_p_values <- p.adjust(raw_p_values, method = "BH")

multiple_testing_results <- data.frame(
  Predictor = names(raw_p_values),
  Raw_P_Value = raw_p_values,
  Adjusted_P_Value = adjusted_p_values
)

multiple_testing_results$Raw_P_Value <- round(multiple_testing_results$Raw_P_Value, 4)
multiple_testing_results$Adjusted_P_Value <- round(multiple_testing_results$Adjusted_P_Value, 4)

multiple_testing_results
```

Multiple p-values were interpreted in the logistic regression model. To reduce the risk of overstating statistical significance, a Benjamini-Hochberg correction was applied. After correction, pregnancies, glucose, and BMI remained statistically significant predictors. Pedigree was significant before correction, but its adjusted p-value was 0.0504, which is slightly above the 0.05 threshold. Therefore, pedigree should be interpreted cautiously as a predictor. Age remained non-significant after correction.