---
title: "Gestational Diabetes Risk Prediction"
author: 'Group 3: Manav, Davis, Daniela, Abenezer'
date: "2026-06-09"
output: word_document
---
##Step 2

# 2a
```{r load/view}
patients <- read.csv("patients.csv")

head(patients)
```

```{r inspect-structure}
str(patients)
```
First, we load the data set into R and inspect the structure of the data set to review the number of observations, variables, variable names, and data types.

```{r summary-stats}
summary(patients)
```
```{r standard-deviation}
sapply(patients, sd)
```
Summary statistics were calculated to examine the spread of each variable. The results showed possible data quality issues because several clinical variables had a minimum of zero, including glucose, blood pressure, skin thickness, insulin, and BMI. These values need to be examined more closely during cleaning steps.

# 2c
```{r loading-ggplot2}
library(ggplot2)
```
```{r}
ggplot(patients, aes(x = factor(Diagnosis))) +
  geom_bar() +
  labs(title = "Count of Diagnosis Groups",
       x = "Diagnosis",
       y = "Count")
```

The bar chart shows that there are more patients without a GEDM diagnosis than with a diagnosis. That means the outcome variable is somewhat imbalanced, which should be considered when evaluating classification model performance.

```{r numeric-features-histogram}
ggplot(patients, aes(x = Glucose)) +
  geom_histogram(bins = 30) +
  labs(title = "Distribution of Glucose",
       x = "Glucose",
       y = "Count")

ggplot(patients, aes(x = BloodPressure)) +
  geom_histogram(bins = 30) +
  labs(title = "Distribution of Blood Pressure",
       x = "Blood Pressure",
       y = "Count")

ggplot(patients, aes(x = BMI)) +
  geom_histogram(bins = 30) +
  labs(title = "Distribution of BMI",
       x = "BMI",
       y = "Count")
```

These histograms show that blood pressure, glucose, and BMI are mostly concentrated around the middle of their ranges. Blood pressure is centered mainly around 60 to 80, glucose around 90 to 130, and BMI around 25 to 40. Each plot also shows a small group of values at or near zero, which stands apart from the rest of the distribution. These zeros will need to be addressed during cleaning.

```{r feature-boxplot-by-diagnosis}
ggplot(patients, aes(x = factor(Diagnosis), y = Pregnancies)) +
  geom_boxplot() +
  labs(title = "Pregnancies by Diagnosis",
       x = "Diagnosis",
       y = "Pregnancies")

ggplot(patients, aes(x = factor(Diagnosis), y = Glucose)) +
  geom_boxplot() +
  labs(title = "Glucose by Diagnosis",
       x = "Diagnosis",
       y = "Glucose")

ggplot(patients, aes(x = factor(Diagnosis), y = BloodPressure)) +
  geom_boxplot() +
  labs(title = "Blood Pressure by Diagnosis",
       x = "Diagnosis",
       y = "Blood Pressure")

ggplot(patients, aes(x = factor(Diagnosis), y = SkinThickness)) +
  geom_boxplot() +
  labs(title = "Skin Thickness by Diagnosis",
       x = "Diagnosis",
       y = "Skin Thickness")

ggplot(patients, aes(x = factor(Diagnosis), y = Insulin)) +
  geom_boxplot() +
  labs(title = "Insulin by Diagnosis",
       x = "Diagnosis",
       y = "Insulin")

ggplot(patients, aes(x = factor(Diagnosis), y = BMI)) +
  geom_boxplot() +
  labs(title = "BMI by Diagnosis",
       x = "Diagnosis",
       y = "BMI")

ggplot(patients, aes(x = factor(Diagnosis), y = Pedigree)) +
  geom_boxplot() +
  labs(title = "Pedigree by Diagnosis",
       x = "Diagnosis",
       y = "Pedigree")

ggplot(patients, aes(x = factor(Diagnosis), y = Age)) +
  geom_boxplot() +
  labs(title = "Age by Diagnosis",
       x = "Diagnosis",
       y = "Age")
```
These box plots compare each predictor between patients without a GDM diagnosis, which is designated as 0, and patients with a GDM diagnosis, which is designated as 1. The diagnosed group generally shows higher values for pregnancies, glucose, BMI, pedigree, and age. Glucose shows one of the clearest separations, with the diagnosed group having noticeably higher median and interquartile range. Blood pressure shows only a small difference between the groups, and most importantly, the data quality issue shown here is that skin thickness and insulin may have values concentrated at zero, which indicate missing values or missing measurements.

```{r age/glucose-diagnosis}
ggplot(patients, aes(x = Age, y = Glucose, color = factor(Diagnosis))) +
  geom_point() +
  labs(title = "Age and Glucose by Diagnosis",
       x = "Age",
       y = "Glucose",
       color = "Diagnosis")
```
This scatterplot shows the relationship between age and glucose, colored with the diagnosis. Diagnosed patients appear more frequently at higher glucose values, while non-diagnosed patients appear to be more concentrated at lower to moderate glucose values. Age does not show as clear of a separation as glucose, because both diagnosis groups appear across much of the age range. The plot also shows a few unusual data points, including glucose values of zero and some unusually high ages, including one around the age of 80, which should be reviewed as possible data quality issues.

```{r glucose/bmi-diagnosis}
ggplot(patients, aes(x = BMI, y = Glucose, color = factor(Diagnosis))) +
  geom_point() +
  labs(title = "BMI and Glucose by Diagnosis",
       x = "BMI",
       y = "Glucose",
       color = "Diagnosis")
```
This scatter plot shows the relationship between BMI and glucose, colored by diagnosis. Diagnosed patients appear more common at higher glucose levels, especially above roughly 125 to 140. BMI also appears somewhat higher among diagnosed patients, but the separation is not as clear. The plot shows possible data quality issues as well, including BMI values at zero and glucose values at zero, which are not clinically realistic and, again, should be reviewed during cleaning.

# 2d
```{r checking-for-na}
colSums(is.na(patients))
```
```{r checking-for-0s}
colSums(patients == 0)
```
No variables contained values coded as NA. However, several variables contained values of zero that are likely invalid or represent missing measurements. Zeros for things like pregnancy and diagnosis are valid because they represent no prior pregnancies and no diagnosis. In contrast, zeros for glucose, blood pressure, skin thickness, insulin, and BMI are not realistic and are highly suspicious, so they should be treated as missing values for cleaning.

```{r convert-to-na}
patients_clean <- patients

invalid_zero_vars <- c("Glucose", "BloodPressure", "SkinThickness", "Insulin", "BMI")

patients_clean[invalid_zero_vars] <- lapply(
  patients_clean[invalid_zero_vars],
  function(x) ifelse(x == 0, NA, x)
)
```
Converting 0s to NA. 

```{r splitting-data}
set.seed(130)

train_index <- unlist(
  tapply(
    1:nrow(patients_clean),
    patients_clean$Diagnosis,
    function(x) sample(x, size = floor(0.70 * length(x)))
  )
)

train_data <- patients_clean[train_index, ]
test_data <- patients_clean[-train_index, ]

nrow(train_data)
nrow(test_data)

table(train_data$Diagnosis)
table(test_data$Diagnosis)
```

Here we are splitting the cleaned patient dataset into a 70/30 training and testing splits based on diagnosis so that both sets maintain a similar proportion of diagnosed to non-diagnosed patients.

```{r}
colSums(is.na(train_data))
colSums(is.na(test_data))
```
```{r}
#define the function to compare median and regression imputation.
compare_imputation_methods <- function(data, target_variable) {
  
#using same seed
  set.seed(130)
  
  predictor_vars <- c("Pregnancies", "Glucose", "BloodPressure", "BMI", "Pedigree", "Age")
  
  predictor_vars <- predictor_vars[predictor_vars != target_variable]
  
#rows where the target variable is actually observed
  observed_rows <- which(!is.na(data[[target_variable]]))
  
#temporarily hiding 30% of observed values for testing imputation accuracy
  hide_rows <- sample(observed_rows, size = round(0.30 * length(observed_rows)))
  
#use the remaining observed rows to build the imputation models.
  train_rows <- setdiff(observed_rows, hide_rows)
  
#temporary copy of the data for this comparison.
  data_for_model <- data
  
#calculate predictor medians using only the comparison training rows.
  predictor_medians <- sapply(
    data_for_model[train_rows, predictor_vars],
    function(x) median(x, na.rm = TRUE)
  )
  
#fill in missing predictor values with training medians.
  for (var in predictor_vars) {
    data_for_model[[var]][is.na(data_for_model[[var]])] <- predictor_medians[[var]]
  }
  
  true_values <- data[[target_variable]][hide_rows]
  
#median-imputed predictions.
  median_value <- median(data[[target_variable]][train_rows], na.rm = TRUE)
  median_predictions <- rep(median_value, length(hide_rows))
  
#the regression formula for the target variable.
  regression_formula <- as.formula(
    paste(target_variable, "~", paste(predictor_vars, collapse = " + "))
  )
  
#fit the regression imputation model.
  regression_model <- lm(
    regression_formula,
    data = data_for_model[train_rows, ]
  )
  
#predict the hidden values using regression imputation.
  regression_predictions <- predict(
    regression_model,
    newdata = data_for_model[hide_rows, ]
  )
  
#error metrics for median and regression imputation.
  data.frame(
    Variable = target_variable,
    Hidden_Values = length(hide_rows),
    Median_MAE = mean(abs(true_values - median_predictions)),
    Regression_MAE = mean(abs(true_values - regression_predictions)),
    Median_RMSE = sqrt(mean((true_values - median_predictions)^2)),
    Regression_RMSE = sqrt(mean((true_values - regression_predictions)^2))
  )
}

rbind(
  compare_imputation_methods(train_data, "Glucose"),
  compare_imputation_methods(train_data, "BloodPressure"),
  compare_imputation_methods(train_data, "BMI"),
  compare_imputation_methods(train_data, "SkinThickness"),
  compare_imputation_methods(train_data, "Insulin")
)
```

Before deciding on the final imputation method, median and regression imputation were compared using the training data. Regression imputation had lower MAE and RMSE for all five variables with missing values, meaning it produced smaller average errors when known values were temporarily hidden and predicted.

```{r imputation-regression}
#copies that will store the imputed data.
train_imputed <- train_data
test_imputed <- test_data

#variables that need imputation.
impute_vars <- c("Glucose", "BloodPressure", "SkinThickness", "Insulin", "BMI")

#predictors used in the imputation models.
impute_predictors <- c("Pregnancies", "Glucose", "BloodPressure", "BMI", "Pedigree", "Age")

#calculate training medians to fill missing predictor values during imputation.
train_medians <- sapply(
  train_data[impute_predictors],
  function(x) median(x, na.rm = TRUE)
)

#function that imputes one target variable at a time.
regression_impute <- function(target_variable) {
  
  predictors <- impute_predictors[impute_predictors != target_variable]
  
#temporary copies for building the imputation model.
  train_for_model <- train_data
  test_for_model <- test_data
  
#fill missing predictor values using training medians.
  for (var in predictors) {
    train_for_model[[var]][is.na(train_for_model[[var]])] <- train_medians[[var]]
    test_for_model[[var]][is.na(test_for_model[[var]])] <- train_medians[[var]]
  }
  
#only training rows where the target variable is observed
  observed_train <- !is.na(train_for_model[[target_variable]])
  
#regression formula for the target variable.
  regression_formula <- as.formula(
    paste(target_variable, "~", paste(predictors, collapse = " + "))
  )
  
#fit the imputation model using training data only.
  impute_model <- lm(
    regression_formula,
    data = train_for_model[observed_train, ]
  )
  
#find missing values in the original train and test sets.
  missing_train <- is.na(train_data[[target_variable]])
  missing_test <- is.na(test_data[[target_variable]])
  
#impute missing training values using the training-based model.
  train_imputed[[target_variable]][missing_train] <<- predict(
    impute_model,
    newdata = train_for_model[missing_train, ]
  )
  
#impute missing testing values using the same training-based model.
  test_imputed[[target_variable]][missing_test] <<- predict(
    impute_model,
    newdata = test_for_model[missing_test, ]
  )
}

#regression imputation on each variable with missing values.
for (var in impute_vars) {
  regression_impute(var)
}

colSums(is.na(train_imputed))
colSums(is.na(test_imputed))
```
Regression imputation was applied as a candidate imputation method because it performed better than median imputation in the error comparison. The imputation models were fit using training data, then applied to both training and testing sets. After imputation, there were no remaining missing values, but the imputed datasets still needed to be checked for impossible or unrealistic values.

```{r}
summary(train_imputed)
summary(test_imputed)
```
Although regression imputation had a lower MAE and RMSE than median, the summary showed that regression imputation produced an impossible negative insulin value. Because insulin cannot possibly be negative, regression imputation was not used as the final imputation method. Median imputation will be selected because it is more stable and does not create impossible clinical values.

```{r impute-median}
#copies that will store the final imputed data
train_imputed <- train_data
test_imputed <- test_data

#variables that need imputation.
impute_vars <- c("Glucose", "BloodPressure", "SkinThickness", "Insulin", "BMI")

#calculate medians from the training data only.
train_medians <- sapply(
  train_data[impute_vars],
  function(x) median(x, na.rm = TRUE)
)

#fill missing values in both sets using training medians.
for (var in impute_vars) {
  train_imputed[[var]][is.na(train_imputed[[var]])] <- train_medians[[var]]
  test_imputed[[var]][is.na(test_imputed[[var]])] <- train_medians[[var]]
}

colSums(is.na(train_imputed))
colSums(is.na(test_imputed))

summary(train_imputed)
summary(test_imputed)
```
Median imputation was used as the final imputation method. The medians were calculated from the training data only and then applied to both the training and testing sets. This avoids using information from the test set while also avoiding impossible imputed values.

#2e
```{r outliers}

outlier_vars <- c("Pregnancies", "Glucose", "BloodPressure", "SkinThickness",
                  "Insulin", "BMI", "Pedigree", "Age")

#IQR bounds using the training data only.
iqr_bounds <- data.frame(
  Variable = outlier_vars,
  Q1 = sapply(train_imputed[outlier_vars], quantile, probs = 0.25, na.rm = TRUE),
  Q3 = sapply(train_imputed[outlier_vars], quantile, probs = 0.75, na.rm = TRUE)
)

iqr_bounds$IQR <- iqr_bounds$Q3 - iqr_bounds$Q1
iqr_bounds$Lower_Bound <- iqr_bounds$Q1 - 1.5 * iqr_bounds$IQR
iqr_bounds$Upper_Bound <- iqr_bounds$Q3 + 1.5 * iqr_bounds$IQR

#count outliers using the training-based IQR bounds.
count_outliers <- function(data, bounds) {
  sapply(bounds$Variable, function(var) {
    lower <- bounds$Lower_Bound[bounds$Variable == var]
    upper <- bounds$Upper_Bound[bounds$Variable == var]
    sum(data[[var]] < lower | data[[var]] > upper)
  })
}

train_outliers <- count_outliers(train_imputed, iqr_bounds)
test_outliers <- count_outliers(test_imputed, iqr_bounds)

iqr_bounds
train_outliers
test_outliers
```
Outliers were identified using the IQR equation after the final median imputation. The IQR bounds were calculated from the training data only and then applied to both training and testing sets. Several variables had outliers, especially insulin, which was expected because insulin had many more missing values, replaced by the training median, creating a narrow IQR. Since these outliers were not clearly impossible values like before, they were handled by retaining them rather than capping or removing.

```{r final-datasets}
train_processed <- train_imputed
test_processed <- test_imputed

summary(train_processed)
summary(test_processed)
```

The final processed training and testing sets created after the outlier review. These sets will be used for feature selection and model building.

##Step 3

#3a

```{r correlation-analysis}
predictor_vars <- c("Pregnancies", "Glucose", "BloodPressure", "SkinThickness",
                    "Insulin", "BMI", "Pedigree", "Age")

#calculate correlations using the training data only.
correlations_with_diagnosis <- data.frame(
  Variable = predictor_vars,
  Correlation_With_Diagnosis = sapply(
    train_processed[predictor_vars],
    function(x) cor(x, train_processed$Diagnosis)
  )
)

#sort variables by strongest absolute correlation with diagnosis.
correlations_with_diagnosis <- correlations_with_diagnosis[
  order(abs(correlations_with_diagnosis$Correlation_With_Diagnosis), decreasing = TRUE),
]

correlations_with_diagnosis
```
This is the correlation analysis which was performed using only the training data. Glucose had the strongest correlation with diagnosis, followed by BMI, pregnancies, and age. All predictors had positive correlation with diagnosis, meaning higher values were generally associated with a higher likelihood of GDM.

```{r stepwise-regression}
#fit full logistic regression model using all predictors.
full_model <- glm(
  Diagnosis ~ Pregnancies + Glucose + BloodPressure + SkinThickness +
    Insulin + BMI + Pedigree + Age,
  data = train_processed,
  family = binomial
)

#use stepwise selection on the training data.
step_model <- step(
  full_model,
  direction = "both",
  trace = 0
)

summary(step_model)
```
The stepwise logistic regression was used for the model-based feature selection on the training data. The stepwise process included pregnancies, glucose, BMI, and pedigree as predictors for the regression model. These variables were selected because they improved the model enough to remain in the final stepwise model, while blood pressure, skin thickness, insulin, and age were not retained. Even though they have some relationship with diagnosis, they did not add enough additional predictive value after the selected values were already included.

```{r selected-variables}
selected_vars <- c("Pregnancies", "Glucose", "BMI", "Pedigree")

selected_vars
```
#3b
```{r standardization}
train_scaled <- train_processed
test_scaled <- test_processed

#calculate means and standard deviations from training data only.
train_means <- sapply(train_processed[selected_vars], mean)
train_sds <- sapply(train_processed[selected_vars], sd)

#standardize selected predictors in the training and testing data.
for (var in selected_vars) {
  train_scaled[[var]] <- (train_processed[[var]] - train_means[[var]]) / train_sds[[var]]
  test_scaled[[var]] <- (test_processed[[var]] - train_means[[var]]) / train_sds[[var]]
}

summary(train_scaled[selected_vars])
summary(test_scaled[selected_vars])
```
The selected predictors were standardized using only the training data. This transformed the selected predictors into z-score type values, where each value represents how many standard deviations it is above or below the training set mean. The same training set means and standard deviations were then applied to the testing data. This approach follows the same preprocessing logic described in the literature review, where preprocessing decisions should be made within the training data to avoid data leakage [16]. Standardization also makes predictors easier to compare because variables such as glucose, BMI, pregnancies, and pedigree are measured on different scales. The final selected predictors are also consistent with the literature, which identifies glucose, BMI, pregnancy history, and family history as important predictors of GDM [4], [9].

##Step 4

#4a

Data partitioning was completed in step 2 before imputation and feature selection. A 70/30 stratified split was used based on diagnosis. This allowed the training and testing sets to maintained similar distribution of diagnosed to non-diagnosed.

#4b
```{r logistic-regression-model}
final_model <- glm(
  Diagnosis ~ Pregnancies + Glucose + BMI + Pedigree,
  data = train_scaled,
  family = binomial
)

summary(final_model)
```
The final logistic regression model uses pregnancies, glucose, BMI, and pedigree as predictors. These four predictors were statistically significant at the 0.05 level, with glucose having the largest coefficient of 1.0669. This aligns with the literature review because glucose measurements are repeatedly identified as central predictors of GDM risk [4], [5]. Pregnancies and BMI were also significant predictors, and this also aligns with the literature, which identifies BMI and pregnancy as relevant risk factors [4], [9]. Pedigree had a smaller but significant effect, which fits its role of family history and genetic predisposition in GDM prediction [4], [9]. Although age was emphasized in the literature, it was not retained by the stepwise model [2], [4], [9]. This suggests that in this dataset, it did not add enough predictive value after the selected variables were included. Overall, this model is reasonably consistent with the literature.

#4c
```{r model-evaluation}
#predicted probabilities for the test set.
test_prob <- predict(
  final_model,
  newdata = test_scaled,
  type = "response"
)

#function to calculate classification metrics at a given threshold.
calculate_metrics <- function(threshold) {
  
  test_pred <- ifelse(test_prob >= threshold, 1, 0)
  
#confusion matrix
  TP <- sum(test_pred == 1 & test_scaled$Diagnosis == 1)
  TN <- sum(test_pred == 0 & test_scaled$Diagnosis == 0)
  FP <- sum(test_pred == 1 & test_scaled$Diagnosis == 0)
  FN <- sum(test_pred == 0 & test_scaled$Diagnosis == 1)
  
#performance metrics from the confusion matrix
  accuracy <- (TP + TN) / (TP + TN + FP + FN)
  precision <- TP / (TP + FP)
  recall <- TP / (TP + FN)
  specificity <- TN / (TN + FP)
  f1 <- 2 * ((precision * recall) / (precision + recall))
  f2 <- 5 * ((precision * recall) / ((4 * precision) + recall))
  
  data.frame(
    Threshold = threshold,
    TP = TP,
    TN = TN,
    FP = FP,
    FN = FN,
    Accuracy = accuracy,
    Precision = precision,
    Recall = recall,
    Specificity = specificity,
    F1 = f1,
    F2 = f2
  )
}

#threshholds 
threshold_results <- do.call(
  rbind,
  lapply(c(0.50, 0.45, 0.35, 0.30), calculate_metrics)
)

threshold_results
```

#4c
```{r model-eval-ROCAUC}
library(pROC)

# Create ROC object using actual diagnoses and predicted probabilities.
roc_model <- roc(
  response = test_scaled$Diagnosis,
  predictor = test_prob,
  quiet = TRUE
)

# Calculate ROC-AUC.
auc_model <- auc(roc_model)
auc_model

# Plot the ROC curve.
plot(
  roc_model,
  main = "ROC Curve for Logistic Regression Model"
)
```

The logistic regression model was evaluated using thresholds of 50%, 45%, 35%, and 30%. Accuracy was very similar across the tested thresholds, ranging from 77.1% to 77.9%.  The threshold decision was based on the balance between precision, recall, specificity, F1, and F2 score. The 35% threshold had the highest accuracy, precision, and specificity, making it a slightly more conservative and balanced option. However, the 30% threshold had the highest recall by far at 84%, the highest F1 score at 72.3%, and the highest F2 score at 78.9%, while maintaining nearly the same accuracy as the 35% threshold. Because this project focuses on identifying patients at risk for GDM, the 30% threshold was selected. This threshold improves the model's ability to identify diagnosed cases, while only slightly reducing precision and specificity, which is appropriate for a screening-oriented prediction model [15].

The ROC AUC for the logistic regression model was 85.8%. This indicates good discrimination between diagnosed and non-diagnosed patients. This suggests that the model is generally effective at ranking higher-risk patients above lower-risk patients. The performance is reasonably consistent with the literature, where GDM prediction models often show useful but variable performance depending on the dataset size, predictors, validation methods, and type of model [4], [5]. Compared with more complex machine learning models, this logistic regression model is simpler and more interpretable, but may be limited by the small dataset and small number of clinical predictors.

#4d
```{r odds-ratios}
#coefficient table from the final logistic regression model.
coef_table <- summary(final_model)$coefficients

#convert log-odds coefficients into odds ratios.
odds_ratios <- data.frame(
  Predictor = rownames(coef_table),
  Estimate = coef_table[, "Estimate"],
  Odds_Ratio = exp(coef_table[, "Estimate"]),
  CI_Lower = exp(coef_table[, "Estimate"] - 1.96 * coef_table[, "Std. Error"]),
  CI_Upper = exp(coef_table[, "Estimate"] + 1.96 * coef_table[, "Std. Error"]),
  P_Value = coef_table[, "Pr(>|z|)"]
)

odds_ratios
```

```{r calibration}
#dataset with actual outcomes and predicted probabilities.
calibration_data <- data.frame(
  Actual = test_scaled$Diagnosis,
  Predicted_Probability = test_prob
)

#spliting patients into 10 groups based on predicted risk.
calibration_data$Risk_Group <- cut(
  rank(calibration_data$Predicted_Probability, ties.method = "first"),
  breaks = 10,
  labels = paste0("Group_", 1:10)
)

#average predicted risk to actual diagnosis rate in each group.
calibration_table <- aggregate(
  cbind(Predicted_Probability, Actual) ~ Risk_Group,
  data = calibration_data,
  FUN = mean
)

names(calibration_table) <- c(
  "Risk_Group",
  "Mean_Predicted_Risk",
  "Observed_GDM_Rate"
)

calibration_table$N <- as.vector(table(calibration_data$Risk_Group))

calibration_table

#Brier score measure for predicted error of probabilites
brier_score <- mean((test_prob - test_scaled$Diagnosis)^2)
brier_score

plot(
  calibration_table$Mean_Predicted_Risk,
  calibration_table$Observed_GDM_Rate,
  xlim = c(0, 1),
  ylim = c(0, 1),
  xlab = "Mean Predicted Risk",
  ylab = "Observed GDM Rate",
  main = "Calibration Plot"
)

abline(0, 1, lty = 2)
```


The model's calibration was assessed by comparing mean predicted risk with observed GDM rate across 10 risk groups. The calibration plot showed that most points followed the reference line fairly closely. This suggests that the model's predicted probabilities were reasonably aligned with actual diagnosis rates. The Brier score was 0.1455, which provides a summary of the probability prediction error.

In addition, the odds ratios aligned with known clinical risk factors for GDM. Because the predictors were standardized, the odds ratio represents the change in odds of diagnosis for one standard deviation increase in that predictor, holding the other predictors consistent. Glucose had the strongest effect, with an odds ratio of 2.91, meaning glucose was associated with substantially higher odds of diagnosis. Pregnancies and BMI also had meaningful positive associations with odds ratios of 1.74 and 1.7. Pedigree had a smaller but still significant effect with an odds ratio of 1.34. This is consistent with the literature, which identifies glucose, BMI, pregnancy history, and family history as important risk factors when trying to predict GDM [4], [5], [9]. 

#4e
```{r multiple-testing-correction}
# Extract p-values for the predictors, excluding the intercept.
predictor_p_values <- summary(final_model)$coefficients[
  rownames(summary(final_model)$coefficients) != "(Intercept)",
  "Pr(>|z|)"
]

# Apply multiple testing corrections to the predictor p-values.
multiple_testing_results <- data.frame(
  Predictor = names(predictor_p_values),
  Unadjusted_P_Value = predictor_p_values,
  BH_Adjusted_P_Value = p.adjust(predictor_p_values, method = "BH"),
  Bonferroni_Adjusted_P_Value = p.adjust(predictor_p_values, method = "bonferroni")
)

multiple_testing_results
```

Multiple testing correction was applied to the p-values for the predictors included in the final logistic regression model. Both Benjamini-Hochberg and Bonferroni corrections were used. All four predictors remained statistically significant after correction. Glucose, pregnancies, and BMI remained highly significant, while pedigree had the weakest adjusted p-value. This supports the final model's predictor-level findings and suggests that selected predictors were not only significant due to multiple testing.

